---
title: "RNA-seq ascites"
author: "Clemens Hug"
date: "2026-01-22"
output:
  html_document:
    theme: flatly
    toc: yes
    number_sections: true
    toc_float:
      collapsed: false
    toc_depth: 4
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(conflicted)
library(tidyverse)
library(data.table)
library(qs)
library(synExtra)
library(DT)
library(powerjoin)
library(here)
library(openxlsx)
library(readxl)
library(DESeq2)
library(ggrepel)
library(ComplexHeatmap)
library(chelpers)
library(seriation)

conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  dplyr::mutate,
  dplyr::rename,
  base::setdiff,
  SummarizedExperiment::rowRanges,
  dplyr::count,
  dplyr::slice,
  purrr::reduce,
  generics::as.factor
)

synapser::synLogin()
syn <- synDownloader("~/data", .cache = TRUE)

dir.create(
  here("ascites"),
  showWarnings = FALSE
)
```

Only "Vehicle" and "YKL" treated mice sequenced for this analysis.

```{r load-data}
counts_raw <- c(
  X2QY7B = "syn72382163",
  F3XVCY = "syn75824014"
) |>
  map(syn) |>
  map(read_tsv)

bridge_sample_ids <- c(3, 15, 18, 32)

meta_raw <- list(
  X2QY7B = syn("syn72382224") |>
    read_excel(
      sheet = "Priority",
      skip = 1L, na = c("", "NA", "N/A")
    ) |>
    select(-Sample...9) |>
    rename(Sample = `Sample...1`) |>
    mutate(
      experiment_batch = "X2QY7B",
      across(Sample, as.character),
      sample_id_seq = seq_len(n()),
      bridge_sample = as.integer(Sample) %in% bridge_sample_ids
    ),
  F3XVCY = syn("syn75824024") |>
    read_csv() |>
    filter(Sample != "BRIDGE SAMPLES") |>
    drop_na(Sample) |>
    mutate(
      bridge_sample = as.integer(Sample) %in% bridge_sample_ids,
      experiment_batch = if_else(
        bridge_sample,
        "X2QY7B",
        "F3XVCY"
      ),
      sample_id_seq = seq_len(n())
    )
)

meta_raw_combined <- bind_rows(
  meta_raw,
  .id = "sequencing_batch"
) |>
  mutate(
    sample_id_unique = paste0(sequencing_batch, "_", sample_id_seq)
  )

View(meta_raw_combined)

write_csv(
  meta_raw_combined,
  here("ascites", "metadata_combined.csv")
)
```

Sample ids for F3XVCY go from 58-87, remapping from 1-34

```{r}
counts_long <- counts_raw |>
  reduce(
    \(left, right) power_full_join(
      left, right,
      by = c("gene_id", "gene_name", "gene_biotype"),
      check = check_specs(
        unmatched_keys_left = "warn",
        unmatched_keys_right = "warn",
        duplicate_keys_left = "warn",
        duplicate_keys_right = "warn"
      )
    )
  ) |>
  pivot_longer(
    cols = -c(gene_id, gene_name, gene_biotype),
    names_to = "sample_id",
    values_to = "count"
  ) |>
  separate_wider_delim(
    sample_id,
    delim = "_",
    names = c("sequencing_batch", "sample_id_seq", "count_type")
  ) |>
  mutate(
    species = case_match(
      str_sub(gene_id, 1, 4),
      "ENSM" ~ "mouse",
      "ENSG" ~ "human",
      .default = NA_character_
    ),
    across(sample_id_seq, as.integer)
  ) |>
  power_inner_join(
    select(meta_raw_combined, sequencing_batch, sample_id_seq, sample_id = Sample, sample_id_unique, bridge_sample),
    by = c("sequencing_batch", "sample_id_seq"),
    check = check_specs(
      unmatched_keys_left = "warn",
      unmatched_keys_right = "warn",
      duplicate_keys_right = "warn"
    )
  )

gene_name_map <- counts_long |>
  distinct(gene_id, gene_name, gene_biotype, species)
```

Remove sample 1, not part of the relevant samples.

```{r}
meta <- meta_raw_combined |>
  filter(
    Group %in% c("Vehicle", "YKL"),
    Sample != 1
  ) |>
  mutate(
    timepoint_factor = recode_values(
      Timepoint,
      0 ~ "initial",
      24 ~ "24h",
      120 ~ "120h",
    ) |>
      fct_relevel("initial"),
    across(Treatment, \(x) replace_na(x, "None") |> fct_relevel("DMSO")),
    across(c(Mouse, sequencing_batch, experiment_batch), \(x) as.factor(as.character(x)))
  )

counts_df <- counts_long |>
  filter(
    count_type == "count"
  ) |>
  select(-count_type)
```

## Total counts per sample

```{r}
p <- ggplot(
  counts_df |>
    drop_na(species) |>
    group_by(sequencing_batch, sample_id, species) |>
    summarise(total_counts = sum(count), .groups = "drop"),
  aes(x = sample_id, y = total_counts, fill = species)
) +
  geom_col(position = "dodge") +
  scale_y_log10(
    expand = expansion(mult = c(0, .05)),
    n.breaks = 8
  ) +
  facet_wrap(
    ~sequencing_batch,
    scales = "free_x"
  ) +
  labs(
    x = "Sample", y = "Total counts"
  ) +
  envalysis::theme_publish()

ggsave(
  here("ascites", "total_counts_per_sample.pdf"),
  p, width = 12, height = 4
)

p
```


## DESeq2 model

We have measurements from multiple mice per group (Vehicle vs. YKL).
Using approach for nested designs from the DESeq2 vignette:

https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#group-specific-condition-effects-individuals-nested-within-groups

```{r}
genes_for_de <- counts_df |>
  group_by(gene_id) |>
  summarise(
    n_samples_expressed = sum(count > 5),
    .groups = "drop"
  ) |>
  filter(
    n_samples_expressed >= 3
  ) |>
  pull(gene_id)

count_mat <- counts_df |>
  filter(gene_id %in% genes_for_de) |>
  select(gene_id, sample_id_unique, count) |>
  pivot_wider(
    names_from = sample_id_unique,
    values_from = count,
    values_fill = 0
  ) |>
  arrange(gene_id) |>
  column_to_rownames("gene_id") |>
  as.matrix()

meta_deseq <- meta |>
  filter(
    Treatment != "None"
  ) |>
  group_by(Group) |>
  mutate(
    # Code mice as 1, 2, ... within each group for nested design
    Mouse_nested = factor(as.integer(factor(Mouse)))
  ) |>
  ungroup() |>
  select(
    sample_id_unique,
    sequencing_batch,
    experiment_batch,
    Group,
    Mouse_nested,
    Treatment,
    timepoint_factor
  )

mm <- model.matrix(
  ~(Group / Mouse_nested) + (Treatment:Group:timepoint_factor),
  data = meta_deseq
)

# remove zero columns
mm <- mm[, colSums(mm) != 0]

qr(mm)$rank
ncol(mm)

de <- DESeqDataSetFromMatrix(
  round(count_mat[, as.character(meta_deseq$sample_id_unique)]),
  colData = meta_deseq |>
    column_to_rownames("sample_id_unique"),
  design = mm
) |>
  DESeq()
```

Getting rank deficient model matrix, trying instead to process two experiments
separately. Making VST counts with simplified model matrix.



### Normalized counts

```{r}
de_combined_simple <- DESeqDataSetFromMatrix(
  round(count_mat[, as.character(meta_deseq$sample_id_unique)]),
  colData = meta_deseq |>
    column_to_rownames("sample_id_unique"),
  design = ~1
) |>
  estimateSizeFactors() |>
  estimateDispersions()

```

```{r}
all_counts <- list(
  raw = counts(de_combined_simple, normalized = FALSE),
  normalized = counts(de_combined_simple, normalized = TRUE),
  vst = assay(vst(de_combined_simple, blind = TRUE))
) |>
  map(\(x) as_tibble(x, rownames = "gene_id")) |>
  bind_rows(.id = "count_type")

write_csv(
  all_counts,
  here("ascites", "all_counts.csv.gz")
)
```

```{r}
des <- meta |>
  filter(
    Treatment != "None",
    sequencing_batch == experiment_batch
  ) |>
  group_by(Group) |>
  mutate(
    # Code mice as 1, 2, ... within each group for nested design
    Mouse_nested = factor(as.integer(factor(Mouse)))
  ) |>
  ungroup() |>
  select(
    sample_id_unique,
    sequencing_batch,
    experiment_batch,
    Group,
    Mouse_nested,
    Treatment,
    timepoint_factor
  ) |>
  group_nest(experiment_batch, timepoint_factor) |>
  mutate(
    mm = map(
      data,
      \(df) {
        mm <- model.matrix(
          ~(Group / Mouse_nested) + (Treatment:Group),
          data = df
        )
        # remove zero columns
        mm <- mm[, colSums(mm) != 0]
        mm
      }
    ),
    de = map2(
      data, mm,
      \(df, mm) DESeqDataSetFromMatrix(
        round(count_mat[, df$sample_id_unique]),
        colData = df |>
          column_to_rownames("sample_id_unique"),
        design = mm
      ) |>
        DESeq()
    )
  )
```



### Extract DESeq2 results

```{r}
extract_deseq_result <- function(de, contrast, name, res_args = list(), shrink_args = list()) {
  if (missing(contrast)) {
    res <- rlang::exec(results, de, name = name, !!!res_args)
  } else {
    res <- rlang::exec(results, de, contrast = contrast, !!!res_args)
  }
  if (!"type" %in% names(shrink_args)) {
    shrink_args$type <- "ashr"
  }
  shrunken <- rlang::exec(lfcShrink, de, res = res, !!!shrink_args)
  shrunken %>%
    as_tibble(rownames = "gene_id") %>%
    left_join(
      res %>%
        as_tibble(rownames = "gene_id") %>%
        select(gene_id, log2FoldChange, lfcSE),
      by = "gene_id", suffix = c("", "_MLE")
    )
}

de_res <- transmute(
  des,
  experiment_batch, timepoint_factor, de,
  comparisons = map(
    de,
    \(de) {
      setdiff(
        resultsNames(de),
        "Intercept"
      ) |>
        str_subset("Group.*\\.Treatment.*") |>
        enframe(value = "covariate") |>
        select(-name) |>
        separate_wider_regex(
          covariate,
          patterns = c(
            "Group", mouse_treatment = "[A-Za-z]+",
            "\\.Treatment", treatment = ".*"
          ),
          cols_remove = FALSE
        ) |>
        mutate(
          across(
            treatment,
            \(x) recode(
              x,
              `RP.6306` = "RP-6306",
              `YKL.Tala` = "YKL+Tala"
            )
          ),
          comparison = paste0("Mouse ", mouse_treatment, " - ", treatment)
        )
    }
  )
) |>
  unnest(comparisons) |>
  mutate(
    de_res = map2(
      de, covariate,
      \(de, covariate) extract_deseq_result(
        de,
        name = covariate
      )
    )
  ) |>
  select(experiment_batch, timepoint_factor, comparison, mouse_treatment, treatment, de_res) |>
  unnest(de_res) |>
  power_left_join(
    gene_name_map,
    by = "gene_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  select(experiment_batch, timepoint_factor, comparison, mouse_treatment, treatment, gene_id, gene_name, gene_biotype, species, everything())

de_res_between_groups <- transmute(
  des,
  experiment_batch, timepoint_factor, de,
  res = map(
    de,
    \(de) {
      meta$Treatment |>
        setdiff(c("DMSO", "None")) |>
        set_names() |>
        map(
          \(treatment) {
            treatment_san <- str_replace_all(treatment, "[^A-Za-z0-9]", ".")
            extract_deseq_result(
              de,
              contrast = list(
                paste0("GroupYKL.Treatment", treatment_san),
                paste0("GroupVehicle.Treatment", treatment_san)
              )
            )
          }
        ) |>
        bind_rows(.id = "treatment") |>
        mutate(
          comparison = paste0("YKL vs Vehicle - ", treatment)
        ) |>
        select(comparison, treatment, gene_id, everything())
    }
  )
) |>
  select(experiment_batch, timepoint_factor, res) |>
  unnest(res) |>
  power_left_join(
    gene_name_map,
    by = "gene_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  select(experiment_batch, timepoint_factor, comparison, treatment, gene_id, gene_name, gene_biotype, species, everything())
```


```{r}
de_res_all <- bind_rows(
  de_res,
  de_res_between_groups
)

write_csv(
  de_res_all,
  here("ascites", "deseq2_results.csv.gz")
)
```

## Volcano plots

```{r}
p <- ggplot(
  de_res |>
    drop_na(padj) |>
    filter(
      gene_biotype == "protein_coding",
      species == "human"
    ),
  aes(
    x = log2FoldChange, y = -log10(padj)
  )
) +
  ggrastr::rasterize(
    geom_point(
      aes(
        color = padj < .05
      ),
      alpha = 0.3,
      shape = 16
    ),
    dpi = 300
  ) +
  geom_text_repel(
    aes(
      label = gene_name
    ),
    data = \(x) group_by(x, comparison, sign(log2FoldChange)) |>
      filter(padj < .05) |>
      slice_min(pvalue, n = 10),
    size = 2,
    max.overlaps = Inf
  ) +
  geom_text(
    aes(
      label = n,
      hjust = if_else(direction == "up", 1, 0),
      x = if_else(direction == "up", Inf, -Inf)
    ),
    data = \(x) filter(
        x, padj < .05
      ) |>
      count(
        comparison,
        direction = if_else(
          log2FoldChange > 0,
          "up",
          "down"
        )
      ),
    y = Inf,
    vjust = 1
  ) +
  scale_color_manual(
    values = c(`FALSE` = "black", `TRUE` = "red"),
    guide = "none"
  ) +
  facet_wrap(
    ~comparison,
    axes = "all"
  ) +
  labs(
    x = "Log2 fold change",
    y = "-Log10 adjusted p-value"
  ) +
  envalysis::theme_publish()
p

ggsave(
  here("ascites", "volcano_plots.pdf"),
  p, width = 12, height = 10
)
```


## PCA plots

```{r}
vst_count_mat <- all_counts |>
  filter(
    count_type == "vst"
  ) |>
  select(-count_type) |>
  column_to_rownames("gene_id") |>
  as.matrix()

pca_raw <- vst_count_mat |>
  t() |>
  prcomp()

pca_sd <- broom::tidy(pca_raw, "eigenvalues")

pc_sd_map <- with(
  pca_sd,
  set_names(percent, paste0("PC", PC))
)

pca_x <- as_tibble(pca_raw$x, rownames = "sample_id_unique") |>
  power_inner_join(
    meta,
    by = c("sample_id_unique"),
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  )

pca_x_long <- as_tibble(pca_raw$x, rownames = "sample_id_unique") |>
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "component",
    values_to = "value"
  )
```


```{r}
pca_x_vs <- combn(paste0("PC", seq_len(6)), 2) |>
  t() |>
  magrittr::set_colnames(paste0("component_", seq_len(2))) |>
  as_tibble() |>
  power_inner_join(
    pca_x_long,
    by = c("component_1" = "component"),
    check = check_specs(
      unmatched_keys_left = "warn"
    )
  ) |>
  power_inner_join(
    pca_x_long,
    by = c("component_2" = "component", "sample_id_unique"),
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    ),
    suffix = c("_1", "_2")
  ) |>
  power_inner_join(
    meta,
    by = c("sample_id_unique"),
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  )

p <- ggplot(
  pca_x_vs,
  aes(
    x = value_1,
    y = value_2,
    color = Treatment,
    shape = Group,
    size = timepoint_factor
  )
) +
  geom_point(alpha = 0.8) +
  scale_size_manual(
    values = c("initial" = 1, "24h" = 2, "120h" = 4)
  ) +
  ggokabeito::scale_color_okabe_ito() +
  # paletteer::scale_color_paletteer_d(
  #   "ggthemes::Tableau_10",
  #   guide = "legend"
  # ) +
  facet_wrap(
    ~paste0(component_1, " (", signif(pc_sd_map[component_1] * 100, 2), "%) vs ", component_2, " (", signif(pc_sd_map[component_2] * 100, 2), "%)"),
    scales = "free"
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  envalysis::theme_publish()

ggsave(
  here("ascites", "pca_plots.pdf"),
  p, width = 12, height = 10
)

p
```

## Clustering

```{r}
cluster_fun <- \(mat) {
  dist_mat <- dist(mat)
  clust <- hclust(dist_mat, method = "ward.D2")
  reorder(clust, dist_mat, method = "OLO")
}

vst_cor_mat <- cor(vst_count_mat)

hm <- Heatmap(
  vst_cor_mat[meta_deseq$sample_id_unique, meta_deseq$sample_id_unique],
  name = "Correlation",
  show_row_names = TRUE,
  show_column_names = FALSE,
  cluster_rows = cluster_fun,
  cluster_columns = cluster_fun,
  top_annotation = create_heatmap_annotation(
    select(
      meta_deseq,
      sample_id_unique, Group, Treatment, timepoint_factor, sequencing_batch
    ) |>
      column_to_rownames("sample_id_unique") |>
      as.data.frame(),
    colors = list(
      Treatment = \(n) paletteer::paletteer_d(
        "ggthemes::Tableau_10", n
      )
    )
  )
)
hm
```


## Mouse YKL vs Vehicle plots

```{r}
ykl_vs_vehicle_de_comp <- de_res |>
  filter(comparison != "Mouse YKL vs Mouse Vehicle") |>
  select(-comparison) |>
  mutate(
    signed_p = -log10(padj) * sign(log2FoldChange)
  ) |>
  pivot_wider(
    names_from = mouse_treatment,
    values_from = c(
      log2FoldChange, lfcSE, pvalue, padj, signed_p, log2FoldChange_MLE, lfcSE_MLE
    )
  ) |>
  power_left_join(
    de_res_between_groups |>
      select(-c(baseMean, gene_name, gene_biotype, species, comparison)) |>
      rename_with(
        .fn = \(x) paste0(x, "_between_groups"),
        .cols = -c(treatment, gene_id)
      ),
    by = c("treatment", "gene_id"),
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn",
      duplicate_keys_left = "warn"
    )
  )

```


```{r}
p <- ggplot(
  ykl_vs_vehicle_de_comp,
  aes(
    x = log2FoldChange_Vehicle,
    y = log2FoldChange_YKL,
    color = padj < .05
  )
) +
  ggrastr::rasterize(geom_point(
    shape = 16, alpha = .6
  ), dpi = 300) +
  scale_color_manual(
    values = c(`FALSE` = "black", `TRUE` = "red")
  ) +
  coord_equal() +
  facet_wrap(
    ~treatment
  ) +
  labs(
    x = "Log2 fold change Vehicle",
    y = "Log2 fold change YKL",
    color = "Adjusted p < 0.05"
  )

ggsave(
  here("ascites", "ykl_vs_vehicle_plots.pdf"),
  p, width = 8, height = 6
)

p
```

## Make Excel sheet

```{r}
format_p <- function(wb, sheet_name, col_name, table) {
  data_rows <- nrow(table) + 1  # +1 for header
  col_idx <- which(names(table) == col_name)

  conditionalFormatting(
    wb = wb,
    sheet = sheet_name,
    cols = col_idx,
    rows = 2:data_rows,  # Skip header row
    type = "expression",
    rule = paste0(
      "AND(NOT(ISBLANK(", openxlsx::int2col(col_idx), "2)),",
      "ISNUMBER(", openxlsx::int2col(col_idx), "2),",
      openxlsx::int2col(col_idx), "2<0.05)"
    ),
    style = createStyle(bgFill = "#FF9999")  # Light red background
  )
}

format_phenotype <- function(wb, sheet_name, col_name, table, percentile = .001) {
  data_rows <- nrow(table) + 1  # +1 for header
  col_idx <- which(names(table) == col_name)

  min_max <- quantile(table[[col_name]], probs = c(percentile, 1 - percentile), na.rm = TRUE) |>
    abs() |>
    max()

  conditionalFormatting(
    wb = wb,
    sheet = sheet_name,
    cols = col_idx,
    rows = 2:data_rows,  # Skip header row
    type = "colourScale",
    style = c("blue", "white", "red"),
    rule = c(-1, 0, 1) * min_max
  )
}

add_sheet <- function(wb, sheet_name, table) {
  addWorksheet(wb, sheet_name)
  table <- select(table, where(negate(is.list)))
  writeDataTable(wb, sheet_name, table, tableStyle = "TableStyleLight1")
  walk(
    str_subset(names(table), "^padj"),
    \(col_name) format_p(wb, sheet_name, col_name, table)
  )
  walk(
    str_subset(names(table), "^log2FoldChange|^signed_p"),
    \(col_name) format_phenotype(wb, sheet_name, col_name, table)
  )
  setColWidths(wb, sheet_name, cols = seq_len(ncol(table)), widths = "auto")
}

wb <- createWorkbook()

list(
  "Treatments" = de_res |>
    arrange(padj, pvalue),
  "Treatments YKL vs Vehicle" = de_res_between_groups |>
    arrange(padj, pvalue),
  "Treatments YKL vs Vehicle Wide" = ykl_vs_vehicle_de_comp |>
    arrange(padj_between_groups, pvalue_between_groups)
) |>
  iwalk(\(t, n) add_sheet(wb, n, t))

# Save the workbook
saveWorkbook(
  wb, here("ascites", "ascites_de_results.xlsx"), overwrite = TRUE
)
```
