---
title: "CDK7 vs GR metrics"
author: "Clemens Hug"
date: "2023-07-28"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)
library(qs)
library(synExtra)

synapser::synLogin()
syn <- synDownloader("~/data", .cache = TRUE)

syn_analysis <- synMkdir("syn52226362", "analysis")
```


```{r}
deseq_res_combined <- synPluck(
  syn_analysis, "cdk7_deseq_res_combined_long.csv"
) %>%
  syn() %>%
  read_csv()

deseq_res_baseline <- synPluck(
  syn_analysis, "cdk7_deseq_baseline_res.csv"
) %>%
  syn() %>%
  read_csv()

counts_baseline <- synPluck(
  syn_analysis, "deep_baseline_counts_long.csv.gz"
) %>%
  syn() %>%
  read_csv() %>%
  mutate(
    cell_line = str_replace_all(cell_line, fixed("-"), "")
  )

```



Comparing gene expression of each cell line compared to the grand
mean of all cell lines at baseline (DMSO)

Because there are big batch effects between cell lines (evidenced)
by the PCA plots, only use the 2019_08 experiment here. That way
we also only have 24h time point instead of 24h and 48h.


```{r read_gr_metrics}
gr_raw <- syn("syn52242760") %>%
  read_csv() %>%
  mutate(neglog_GR50 = -log10(GR50))

```


```{r}
p <- de_baseline_res %>%
  ggplot(
    aes(log2FoldChange, -log10(padj))
  ) +
  ggrastr::rasterize(geom_point(shape = 16, alpha = 0.3), dev = "ragg", dpi = 200) +
  facet_wrap(~cell_line)

ggsave(
  file.path("plots_cdk7", "baseline_lfc_padj.pdf"),
  p,
  width = 10,
  height = 8
)
```


```{r}
de_baseline_n_sig <- de_baseline_res %>%
  group_by(cell_line) %>%
  summarize(
    n_sig = sum(padj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

p <- de_baseline_n_sig %>%
  ggplot(
    aes(cell_line, n_sig)
  ) +
    geom_col() +
    coord_flip() +
    labs(x = NULL, y = "Number of significant genes")

ggsave(
  file.path("plots_cdk7", "baseline_n_sig.pdf"),
  p,
  width = 6,
  height = 4
)

de_baseline_selected_genes <- de_baseline_res %>%
  group_by(ensembl_gene_id) %>%
  filter(any(padj < 0.05)) %>%
  ungroup() %>%
  pull(ensembl_gene_id) %>%
  unique()
```

Correlate GR50 vs log2FC of each cell line compared to the grand
mean. Hopefully we're able to identify some genes that are correlated

```{r}
baseline_lfc_mat <- de_baseline_res %>%
  select(cell_line, ensembl_gene_id, log2FoldChange) %>%
  pivot_wider(names_from = cell_line, values_from = log2FoldChange, values_fill = 0) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()


gr_raw %>%
  filter(cell_line %in% baseline_meta$cell_line) %>%
  View()

gr_mats <- gr_raw %>%
  select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
  column_to_rownames("cell_line") %>%
  as.matrix() %>%
  t() %>% {
    .[, baseline_meta$cell_line]
  }

baseline_lfc_mat_overlap <- baseline_lfc_mat[
  de_baseline_selected_genes, baseline_meta$cell_line
]

compute_correlation <- function(gr_mat, lfc_mat, ...) {
  raw_cor <- psych::corr.test(
    t(lfc_mat),
    t(gr_mat),
    adjust = "fdr",
    ci = FALSE,
    ...
  )
  cor_df <- raw_cor[c("r", "p", "p.adj")] %>%
    imap(
      ~as_tibble(.x, rownames = "ensembl_gene_id") %>%
        pivot_longer(
          cols = -ensembl_gene_id,
          names_to = "gr_metric",
          values_to = .y
        )
    ) %>%
    purrr::reduce(
      left_join,
      by = c("ensembl_gene_id", "gr_metric")
    ) %>%
    # left_join(
    #   ensembl_gtf %>%
    #     select(ensembl_gene_id, hgnc_symbol, gene_name),
    #   by = "ensembl_gene_id"
    # ) %>%
    group_by(gr_metric) %>%
    mutate(
      p.adj = p.adjust(
        p, method = "BH"
      )
    ) %>%
    ungroup()
  cor_df
}

baseline_lfc_cor_df <- compute_correlation(
  gr_mats,
  baseline_lfc_mat_overlap
)

p <- baseline_lfc_cor_df %>%
  pivot_longer(
    cols = -c(ensembl_gene_id, hgnc_symbol, gene_name, gr_metric),
    names_to = "cor_metric",
    values_to = "value"
  ) %>%
  ggplot(aes(value, color = cor_metric)) +
    stat_ecdf(geom = "step") +
    facet_wrap(~gr_metric)

dir.create("plots_cdk7")
ggsave(
  file.path("plots_cdk7", "baseline_lfc_cor_ecdf.pdf"),
  p,
  width = 10,
  height = 8
)

p <- baseline_lfc_cor_df %>%
  select(-lfc_var) %>%
  pivot_longer(
    cols = -c(ensembl_gene_id, hgnc_symbol, gene_name, gr_metric),
    names_to = "cor_metric",
    values_to = "value"
  ) %>%
  ggplot(aes(value, color = cor_metric)) +
    geom_freqpoly() +
    facet_wrap(~gr_metric)

baseline_lfc_cor_sig <- baseline_lfc_cor_df %>%
  group_by(gr_metric) %>%
  arrange(p) %>%
  slice_head(n = 10) %>%
  ungroup()
  # filter(p.adj < 0.1) %>%
  # arrange(p.adj)

p <- baseline_lfc_cor_df %>%
  inner_join(
    de_baseline_res
  ) %>%
  inner_join(
    gr_raw %>%
      select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
      pivot_longer(-cell_line, names_to = "gr_metric", values_to = "gr_value"),
  ) %>%
  ggplot(
    aes(gr_value, log2FoldChange, color = gr_metric)
  ) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~gene_name, scales = "free")

p <- baseline_lfc_cor_df %>%
  filter(gr_metric %in% c("neglog_GR50", "GRmax")) %>%
  filter(
    ensembl_gene_id %in% {
      group_by(., gr_metric) %>%
        arrange(p) %>%
        slice_head(n = 10) %>%
        pull(ensembl_gene_id)
    }
  ) %>%
  inner_join(
    de_baseline_res
  ) %>%
  inner_join(
    gr_raw %>%
      select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
      pivot_longer(-cell_line, names_to = "gr_metric", values_to = "gr_value"),
  ) %>%
  group_nest(gr_metric) %>%
  mutate(
    p = map2(
      data, gr_metric,
      function(df, g) {
        ggplot(
          df %>%
            arrange(p) %>%
            mutate(across(gene_name, fct_inorder)),
          aes(gr_value, log2FoldChange)
        ) +
          geom_point() +
          geom_smooth(method = "lm") +
          geom_text(
            aes(x = x, y = y, label = p),
            hjust = 0, vjust = 1,
            nudge_x = 0.1, nudge_y = -0.1,
            data = \(df) distinct(df, gene_name, p) %>%
              mutate(
                across(p, \(x) signif(x, digits = 2)),
                x = if (g == "GR50") Inf else -Inf, y = Inf
              )
          ) +
          facet_wrap(~gene_name, scales = "free") +
          scale_x_continuous(
            trans = switch(
              g,
              neglog_GR50 = "identity",
              GR50 = c("reverse"),
              "identity"
            )
          ) +
          labs(
            x = g,
            y = "log2FoldChange cell line vs grand mean",
            title = g
          )
      }
    )
  )

ps <- patchwork::wrap_plots(p$p)

ggsave(
  file.path("plots_cdk7", "baseline_lfc_cor_most_sig.pdf"),
  ps,
  width = 16,
  height = 8
)

p <- baseline_lfc_cor_df %>%
  ggplot(
    aes(cor, color = metric)
  ) +
    stat_ecdf(geom = "step")

baseline_lfc_cor_highest <- baseline_lfc_cor_df %>%
  arrange(cor) %>%
  group_by(metric) %>%
  dplyr::slice(
    1:10, (n() - 9):n()
  ) %>%
  ungroup() %>%
  left_join(
    ensembl_gtf %>%
      select(ensembl_gene_id, hgnc_symbol),
    by = "ensembl_gene_id"
  )
```

## Baseline deep counts


```{r}


gr_long <- gr_raw %>%
  mutate(
    cell_line = str_replace_all(cell_line, fixed("-"), "")
  ) %>%
  select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
  pivot_longer(
    cols = -cell_line,
    names_to = "gr_metric",
    values_to = "gr_value"
  )

gr_mats <- gr_raw %>%
  mutate(
    cell_line = str_replace_all(cell_line, fixed("-"), "")
  ) %>%
  select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
  column_to_rownames("cell_line") %>%
  as.matrix() %>%
  t()

intersect(
  counts_baseline$cell_line,
  colnames(gr_mats)
)
setdiff(
  counts_baseline$cell_line,
  colnames(gr_mats)
)
setdiff(
  colnames(gr_mats),
  counts_baseline$cell_line
)
deep_baseline_cell_line_overlap <- intersect(
  counts_baseline$cell_line,
  colnames(gr_mats)
)

deep_baseline_eligible_genes <- counts_baseline %>%
  filter(
    type == "raw",
    cell_line %in% deep_baseline_cell_line_overlap
  ) %>%
  group_by(ensembl_gene_id) %>%
  summarize(
    n_above_10 = sum(count > 10), .groups = "drop"
  ) %>%
  filter(n_above_10 > 5) %>%
  pull(ensembl_gene_id)

baseline_deep_varstab_mat <- counts_baseline %>%
  filter(
    type == "varstab",
    cell_line %in% deep_baseline_cell_line_overlap,
    ensembl_gene_id %in% deep_baseline_eligible_genes
  ) %>%
  select(cell_line, ensembl_gene_id, count) %>%
  pivot_wider(names_from = cell_line, values_from = count, values_fill = 0) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix() %>%
  {.[, deep_baseline_cell_line_overlap]}

gr_mats_ordered <- gr_mats[, deep_baseline_cell_line_overlap]

baseline_deep_varstab_cor_df <- compute_correlation(
  gr_mats_ordered,
  baseline_deep_varstab_mat,
  # log10(baseline_deep_varstab_mat + .5 * min(baseline_deep_varstab_mat[baseline_deep_varstab_mat > 0])),
  method = "pearson"
) %>%
  left_join(
    counts_baseline %>%
      distinct(ensembl_gene_id, hgnc_symbol, gene_name),
    by = "ensembl_gene_id"
  )
```

```{r}
library(ggrepel)

p <- baseline_deep_varstab_cor_df %>%
  drop_na(r) %>%
  ggplot(
    aes(r, -log10(p))
  ) +
  ggrastr::rasterize(geom_point(), dev = "ragg", dpi = 150) +
  geom_text_repel(
    aes(label = gene_name),
    data = \(x) group_by(
      x, gr_metric
    ) %>%
      arrange(p) %>%
      mutate(
        gene_name = if_else(
          row_number() %in% 1:5,
          gene_name,
          ""
        )
      ),
    max.overlaps = Inf
  ) +
  ggh4x::facet_wrap2(
    ~gr_metric
  )
```


```{r}
baseline_deep_varstab_cor_df_top_10_data <- baseline_deep_varstab_cor_df %>%
  arrange(p) %>%
  group_by(gr_metric) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  left_join(
    counts_baseline %>%
      filter(type == "normalized", cell_line %in% deep_baseline_cell_line_overlap),
    by = c("ensembl_gene_id", "hgnc_symbol", "gene_name")
  ) %>%
  powerjoin::power_left_join(
    gr_long,
    by = c("cell_line", "gr_metric"),
    check = powerjoin::check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  )

ps <- baseline_deep_varstab_cor_df_top_10_data %>%
  group_nest(gr_metric) %>%
  rowwise() %>%
  mutate(
    p = list(
      ggplot(
        data %>%
          mutate(count = count + .5 * min(count[count > 0])),
        aes(count, gr_value, color = cell_line)
      ) +
        geom_point() +
        geom_smooth(color = "black", method = "lm") +
        facet_wrap(~gene_name, scales = "free") +
        labs(
          y = gr_metric,
          x = "Variance stabilized count",
          title = gr_metric
        ) +
        scale_x_log10() +
        theme_linedraw()
    )
  )

pwalk(
  ps,
  \(gr_metric, p, ...) {
    ggsave(
      file.path("plots_cdk7", paste0("deep_baseline_vs_gr_top10_correlated_", gr_metric, ".pdf")),
      p, width = 12, height = 8
    )
  }
)
```


```{r}
library(fgsea)
library(msigdbr)

all_gene_sets <- msigdbr(species = "Homo sapiens")
selected_gene_sets <- all_gene_sets %>%
  filter(
    gs_cat == "H" | (
      gs_subcat == "CP:KEGG"
    ) | gs_subcat == "CP:REACTOME"
  )

selected_gene_sets_gr <- selected_gene_sets %>%
  group_nest(gs_cat, gs_subcat, gs_name) %>%
  mutate(
    gs = map(
      data,
      \(x) unique(x$ensembl_gene)
    ) %>%
      set_names(gs_name)
  )

selected_des <- baseline_deep_varstab_cor_df %>%
  drop_na(r) %>%
  group_nest(gr_metric) %>%
  mutate(
    de = map(
      data,
      \(x) set_names(x$r, x$ensembl_gene_id)
    ),
    de_p = map(
      data,
      \(x) set_names(-sign(x$r) * log10(x$p), x$ensembl_gene_id)
    )
  )

fgsea_res_raw <- selected_des %>%
  mutate(
    res = map(
      de,
      \(x) fgseaMultilevel(
        selected_gene_sets_gr$gs,
        x
      )
    ),
    res_p = map(
      de_p,
      \(x) fgseaMultilevel(
        selected_gene_sets_gr$gs,
        x
      )
    )
  )

fgsea_res_p_long <- fgsea_res_raw %>%
  select(gr_metric, res_p) %>%
  unnest(res_p) %>%
  mutate(
    signed_p = -sign(NES) * log10(padj),
    p_cut = cut(padj, breaks = c(-Inf, 0.01, 0.05, Inf), labels = c("**", "*", ""))
  ) %>%
  filter(
    pathway %in% {
      group_by(., pathway) %>%
        summarize(n_sig = sum(padj < 0.05)) %>%
        filter(n_sig > 0) %>%
        pull(pathway)
    }
  )

ps <- fgsea_res_p_long %>%
  filter(padj < .1) %>%
  group_nest(
    gr_metric
  ) %>%
  rowwise() %>%
  mutate(
    p = list(
      ggplot(
        data %>%
          arrange(NES) %>%
          mutate(
            pathway = fct_inorder(pathway)
          ),
        aes(NES, pathway, fill = p_cut)
      ) +
        geom_col() +
        geom_text(
          aes(
            x = x_pos,
            y = pathway,
            label = pathway,
            hjust = hjust
          ),
          data = \(x) mutate(
            x,
            x_pos = if_else(NES > 0, -.1, .1),
            hjust = if_else(NES > 0, 1, 0)
          ),
          inherit.aes = FALSE
        ) +
        scale_fill_viridis_d(direction = -1) +
        labs(
          x = "Normalized Enrichment Score",
          y = NULL,
          title = gr_metric
        ) +
        theme(
          axis.text.y = element_blank()
        )
    )
  )

pwalk(
  ps,
  \(gr_metric, p, ...) {
    ggsave(
      file.path("plots_cdk7", paste0("fgsea_bars_", gr_metric, ".pdf")),
      p, width = 8, height = 12
    )
  }
)
```

```{r}

library(seriation)
cluster_df <- function(df, row_var, col_var, value_var, values_fill = 0) {
  # browser()
  mat <- df %>%
    select({{row_var}}, {{col_var}}, {{value_var}}) %>%
    pivot_wider(names_from = {{col_var}}, values_from = {{value_var}}, values_fill = values_fill) %>%
    column_to_rownames(rlang::as_name(rlang::enquo(row_var)))
  # browser()
  if (rlang::is_bare_numeric(pull(df, {{value_var}}))) {
    dist_rows <- dist(mat, method = "euclidian")
    dist_cols <- dist(t(mat), method = "euclidian")
  } else {
    # browser()
    dist_rows <- cluster::daisy(mat, metric = "gower")
    dist_cols <- t(mat) %>%
      as.data.frame() %>%
      mutate(across(everything(), \(x) factor(x, levels = levels(pull(df, {{value_var}}))))) %>%
      cluster::daisy(metric = "gower")
  }
  clust_rows <- hclust(dist_rows, method = "average") %>%
      reorder(dist_rows, method = "olo")
  clust_cols <- hclust(dist_cols, method = "average") %>%
      reorder(dist_cols, method = "olo")
  df %>%
    mutate(
      "{{row_var}}" := factor({{row_var}}, levels = clust_rows$labels[clust_rows$order]),
      "{{col_var}}" := factor({{col_var}}, levels = clust_cols$labels[clust_cols$order])
    )
}

better_contrast <- function(
  colors1, colors2, threshold = 0
) {
  # browser()
  farver::compare_colour(
    farver::decode_colour(colors1),
    farver::decode_colour(colors2),
    from_space = "rgb",
    method = "cie2000"
  ) %>% {
      .[, 1] < (1 + threshold) * .[, 2]
    } %>%
    if_else(colors2[2], colors2[1])
}

p <- fgsea_res_p_long %>%
  filter(
    gr_metric != "GR50",
    str_starts(pathway, "REACTOME")
  ) %>%
  filter(
    pathway %in% {
      group_by(., pathway) %>%
        summarize(n_sig = sum(padj < 0.01)) %>%
        filter(n_sig > 0) %>%
        pull(pathway)
    }
  ) %>%
  cluster_df(
    pathway, gr_metric, NES
  ) %>%
  ggplot(
    aes(gr_metric, pathway, fill = NES)
  ) +
    geom_raster() +
    geom_text(
      aes(
        label = p_cut,
        color = after_scale(better_contrast(fill, c("white", "black"), 0))
      ),
      hjust = .5, vjust = .5
    ) +
    scico::scale_fill_scico(
      palette = "vik", midpoint = 0
    ) +
    theme_minimal()

```


```{r}
ykl_cell_lines_to_keep <- intersect(
  deseq_res_long %>%
    filter(agent == "YKL-5-124", experiment == "cdk_467_2019_08") %>%
    pull(cell_line),
  gr_raw$cell_line
)

ykl_genes_to_keep <- deseq_res_long %>%
  filter(
    agent == "YKL-5-124",
    experiment == "cdk_467_2019_08",
    cell_line %in% ykl_cell_lines_to_keep
  ) %>%
  group_by(ensembl_gene_id) %>%
  filter(any(padj < 0.05)) %>%
  ungroup() %>%
  pull(ensembl_gene_id) %>%
  unique()

ykl_lfc_mat <- deseq_res_long %>%
  filter(
    agent == "YKL-5-124",
    experiment == "cdk_467_2019_08",
    cell_line %in% ykl_cell_lines_to_keep,
    ensembl_gene_id %in% ykl_genes_to_keep
  ) %>%
  select(ensembl_gene_id, cell_line, log2FoldChange) %>%
  replace_na(list(log2FoldChange = 0)) %>%
  pivot_wider(names_from = cell_line, values_from = log2FoldChange, values_fill = 0) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()

ykl_gr_mat <- gr_raw %>%
  filter(cell_line %in% ykl_cell_lines_to_keep) %>%
  select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
  column_to_rownames("cell_line") %>%
  as.matrix() %>%
  t()

ykl_gr_cor_df <- compute_correlation(
  ykl_gr_mat,
  ykl_lfc_mat
)

p <- ykl_gr_cor_df %>%
  pivot_longer(
    cols = c(r, p, p.adj),
    names_to = "cor_metric",
    values_to = "value"
  ) %>%
  ggplot(aes(value, color = cor_metric)) +
    stat_ecdf(geom = "step") +
    facet_wrap(~gr_metric)

ggsave(
  file.path("plots_cdk7", "ykl_gr_cor_ecdf.pdf"),
  p,
  width = 10,
  height = 8
)

```

```{r}
ykl_gr_cor_sig <- ykl_gr_cor_df %>%
  filter(gr_metric %in% c("GR50", "neglog_GR50")) %>%
  filter(
    ensembl_gene_id %in% {
      group_by(., gr_metric) %>%
        arrange(p) %>%
        slice_head(n = 10) %>%
        pull(ensembl_gene_id)
    }
  )


ykl_gr_cor_sig_lfc <- ykl_gr_cor_sig %>%
  inner_join(
    ykl_lfc_mat %>%
      as_tibble(rownames = "ensembl_gene_id") %>%
      pivot_longer(
        cols = -ensembl_gene_id,
        names_to = "cell_line",
        values_to = "log2FoldChange"
      )
  ) %>%
  inner_join(
    gr_raw %>%
      select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
      pivot_longer(-cell_line, names_to = "gr_metric", values_to = "gr_value"),
  )

p <- ykl_gr_cor_sig_lfc %>%
  group_nest(gr_metric) %>%
  mutate(
    p = map2(
      data, gr_metric,
      function(df, g) {
        ggplot(
          df %>%
            arrange(p) %>%
            mutate(across(gene_name, fct_inorder)),
          aes(gr_value, log2FoldChange)
        ) +
          geom_point() +
          geom_smooth(method = "lm") +
          geom_text(
            aes(x = x, y = y, label = p),
            hjust = 0, vjust = 1,
            nudge_x = 0.1, nudge_y = -0.1,
            data = \(df) distinct(df, gene_name, p) %>%
              mutate(
                across(p, \(x) signif(x, digits = 2)),
                x = if (g == "GR50") Inf else -Inf, y = Inf
              )
          ) +
          facet_wrap(~gene_name, scales = "free") +
          scale_x_continuous(
            trans = switch(
              g,
              neglog_GR50 = "identity",
              GR50 = c("reverse"),
              "identity"
            )
          ) +
          labs(
            x = g,
            y = "log2FoldChange YKL-5-124 vs DMSO",
            title = g
          )
      }
    )
  )

ps <- patchwork::wrap_plots(p$p)

ggsave(
  file.path("plots_cdk7", "ykl_gr_cor_most_sig.pdf"),
  ps,
  width = 16,
  height = 8
)

```

```{r}
cv_model <- glmnet::cv.glmnet(
  t(ykl_lfc_mat), ykl_gr_mat["neglog_GR50",],
  alpha = 0.5,
  # nfolds = ncol(ykl_lfc_mat),
  foldid = 1:ncol(ykl_lfc_mat)
)

best_model <- glmnet::glmnet(
  t(ykl_lfc_mat), ykl_gr_mat["neglog_GR50",],
  alpha = 0.5, lambda = cv_model$lambda.min
)

sparse_mat_to_df <- function(x) {
  y <- summary(as(x, "dgTMatrix"))
  y$row = rownames(x)[y$i]
  y$col = colnames(x)[y$j]
  y
}

best_model_coef <- sparse_mat_to_df(coef(best_model)) %>%
  dplyr::rename(
    ensembl_gene_id = row,
    coef = x
  ) %>%
  left_join(
    ensembl_gtf %>%
      select(ensembl_gene_id, hgnc_symbol, gene_name),
    by = "ensembl_gene_id"
  ) %>%
  as_tibble()

best_model_coef_lfc <- best_model_coef %>%
  mutate(gr_metric = "neglog_GR50") %>%
  inner_join(
    ykl_lfc_mat %>%
      as_tibble(rownames = "ensembl_gene_id") %>%
      pivot_longer(
        cols = -ensembl_gene_id,
        names_to = "cell_line",
        values_to = "log2FoldChange"
      )
  ) %>%
  inner_join(
    gr_raw %>%
      select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
      pivot_longer(-cell_line, names_to = "gr_metric", values_to = "gr_value"),
  )

p <- best_model_coef_lfc %>%
  group_nest(gr_metric) %>%
  mutate(
    p = map2(
      data, gr_metric,
      function(df, g) {
        ggplot(
          df %>%
            arrange(coef) %>%
            mutate(across(gene_name, fct_inorder)),
          aes(gr_value, log2FoldChange)
        ) +
          geom_point() +
          geom_smooth(method = "lm") +
          geom_text(
            aes(x = x, y = y, label = coef),
            hjust = 0, vjust = 1,
            nudge_x = 0.1, nudge_y = -0.1,
            data = \(df) distinct(df, gene_name, coef) %>%
              mutate(
                across(coef, \(x) signif(x, digits = 2)),
                x = if (g == "GR50") Inf else -Inf, y = Inf
              )
          ) +
          facet_wrap(~gene_name, scales = "free") +
          scale_x_continuous(
            trans = switch(
              g,
              neglog_GR50 = "identity",
              GR50 = c("reverse"),
              "identity"
            )
          ) +
          labs(
            x = g,
            y = "log2FoldChange YKL-5-124 vs DMSO",
            title = g
          )
      }
    )
  )

ps <- patchwork::wrap_plots(p$p)

ggsave(
  file.path("plots_cdk7", "ykl_gr_lasso_most_sig.pdf"),
  ps,
  width = 8,
  height = 8
)

```


```{r}
model_res_table <- list(
  baseline = baseline_lfc_cor_df %>%
    select(
      ensembl_gene_id, hgnc_symbol,
      gr_metric, r, p, p.adj
    ),
  ykl = ykl_gr_cor_df %>%
    select(
      ensembl_gene_id, hgnc_symbol,
      gr_metric, r, p, p.adj
    ),
  ykl_elastic_net = best_model_coef %>%
    transmute(
      ensembl_gene_id, hgnc_symbol,
      gr_metric = "neglog_GR50",
      coef
    )
) %>%
  purrr::reduce2(
    .y = names(.),
    .init = tibble(ensembl_gene_id = character(), hgnc_symbol = character(), gr_metric = character()),
    \(df1, df2, prefix) {
      full_join(
        df1,
        rename_with(df2, ~paste0(prefix, "_", .x), -c(ensembl_gene_id, hgnc_symbol, gr_metric)),
        by = c("ensembl_gene_id", "hgnc_symbol", "gr_metric")
      )
    }
  ) %>%
  replace_na(list(ykl_elastic_net_coef = 0)) %>%
  group_by(gr_metric) %>%
  mutate(
    rank_ykl_elastic_net = rank(-abs(ykl_elastic_net_coef)),
    rank_ykl = rank(ykl_p),
    rank_baseline = rank(baseline_p),
    rank_average = (rank_ykl_elastic_net + rank_ykl) / 3
  ) %>%
  ungroup() %>%
  mutate(
    ensembl_gene_id = factor(
      ensembl_gene_id,
      levels = filter(., gr_metric %in% c("neglog_GR50")) %>%
        group_by(ensembl_gene_id) %>%
        summarize(
          rank_average = mean(rank_average),
          .groups = "drop"
        ) %>%
        arrange(rank_average) %>%
        pull(ensembl_gene_id)
    )
  ) %>%
  arrange(ensembl_gene_id) %>%
  filter(gr_metric %in% c("neglog_GR50", "GR50"))

xlsx::write.xlsx(
  model_res_table,
  file.path("plots_cdk7", "gr_correlation_table.xlsx")
)

```

