---
title: "DESeq"
author: "Clemens Hug"
date: "2023-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)
library(qs)
library(DESeq2)
```

Reading count and metadata tables generated in `qc.Rmd`.

```{r}
de_meta <- read_csv("deseq/de_meta.csv.gz")
de_counts <- read_csv("deseq/de_counts.csv.gz")

de_count_mat <- de_counts %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()
```

Only consider genes that have at least 5 counts in more than 10 samples to
speed up analysis. Doesn't have impact on outcome because DESeq2 would exclude
these anyways, but take longer.

```{r}
genes_to_keep <- de_count_mat %>%
  magrittr::is_greater_than(5) %>%
  rowSums() %>% {
    names(.)[. >= 10]
  }
```

Perform differential expression analysis by plate. No treatment / control pairs
cross plate boundaries so splitting it analysis up by plate makes things easier
and faster here.

I chose to set up a single model for each plate, instead of separate models for
each treatment / control pair, because estimates of expression variance might
be better that way, see https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#if-i-have-multiple-groups-should-i-run-all-together-or-split-into-pairs-of-groups

```{r}
de_groups <- de_meta %>%
  group_by(source_96) %>%
  summarize(
    des = DESeqDataSetFromMatrix(
      de_count_mat[genes_to_keep, well],
      cur_data(),
      ~condition
    ) %>%
      DESeq(parallel = TRUE, BPPARAM = BiocParallel::MulticoreParam(workers = 6)) %>%
      list()
  )
```

Identifying controls. Anything with DMSO agent is control.

```{r}
de_controls <- de_meta %>%
  filter(agent == "DMSO") %>%
  distinct(
    source_96, cell_line, timepoint, control_condition = condition
  )
```

Matching treatments with appropriate controls by checking whether for any given
treatment we have DMSO control with matching plate, cell_line, and time point.

```{r}
de_comps <- de_meta %>%
  filter(agent != "DMSO") %>%
  distinct(
    source_96, cell_line, agent, concentration, timepoint, condition
  ) %>%
  inner_join(
    de_controls
  )

de_results <- de_groups %>%
  inner_join(
    de_comps
  ) %>%
  rowwise() %>%
  mutate(
    de_res = results(
      des,
      contrast = c("condition", condition, control_condition),
      tidy = TRUE,
      parallel = TRUE, BPPARAM = BiocParallel::MulticoreParam(workers = 6)
    ) %>%
      dplyr::rename(ensembl_gene_id = row) %>%
      list()
  ) %>%
  ungroup()
```

Compute number of differentially expressed genes per treatment condition and
average number of gene counts in the contributing wells.

```{r}
well_counts <- de_count_mat %>%
  colSums() %>%
  enframe("well", "count") %>%
  inner_join(
    de_meta %>%
      distinct(well, condition),
    by = "well"
  )

de_stats <- de_results %>%
  select(-des) %>%
  mutate(
    n_de = map_int(de_res, ~nrow(filter(.x, padj < 0.05)))
  ) %>%
  select(-de_res) %>%
  rowwise() %>%
  mutate(
    count_avg = mean(well_counts[well_counts$condition %in% c(condition, control_condition), ]$count)
  ) %>%
  ungroup()

```

Visualize number of DE genes using beeswarm plots.

```{r}
library(ggbeeswarm)
library(ggrepel)

p <- de_stats %>%
  ggplot(aes(cell_line, n_de, color = source_96)) +
    geom_quasirandom() +
    geom_text_repel(
      aes(label = agent),
      show.legend = FALSE,
      data = ~.x %>%
        group_by(cell_line) %>%
        arrange(desc(n_de)) %>%
        mutate(agent = if_else(seq_len(n()) <= 4, agent, "")) %>%
        ungroup()
    ) +
    labs(color = "Plate", y = "N differentially expressed", x = "Cell line")

dir.create("plots/deseq")
ggsave(
  "plots/deseq/n_de_beeswarm.pdf", p, width = 7, height = 6
)

p <- de_stats %>%
  ggplot(aes(source_96, n_de, color = cell_line)) +
    geom_quasirandom() +
    geom_text_repel(
      aes(label = agent, color = NULL),
      show.legend = FALSE,
      box.padding = 0.5,
      max.overlaps = Inf,
      data = ~.x %>%
        group_by(source_96) %>%
        arrange(desc(n_de)) %>%
        mutate(agent = if_else(seq_len(n()) <= 4, agent, "")) %>%
        ungroup()
    ) +
    labs(x = "Plate", y = "N differentially expressed", color = "Cell line")

ggsave(
  "plots/deseq/n_de_beeswarm_by_plate.pdf", p, width = 7, height = 6
)
```


```{r}
p <- de_stats %>%
  ggplot(aes(count_avg, n_de, color = cell_line, shape = source_96)) +
    geom_point() +
    labs(x = "Mean well count", y = "N differentially expressed", color = "Cell line", shape = "Plate")

ggsave(
  "plots/deseq/n_de_vs_well_count.pdf", p, width = 7, height = 6
)

```

Perform differential expression for matching samples from old DGE experiments

```{r old_datasets}
old_samples_de_meta <- fread("deseq/old_samples_de_meta.csv.gz")
old_samples_de_counts <- fread("deseq/old_samples_raw_counts_long.csv.gz")

old_samples_de_count_mat <- old_samples_de_counts %>%
  

de_old_groups <- old_samples_de_meta %>%
  group_by(experiment) %>%
  summarize(
    des = DESeqDataSetFromMatrix(
      old_samples_de_counts[old_samples_de_meta$well %in% well & old_samples_de_meta$experiment == experiment[1]]
      de_count_mat[genes_to_keep, well],
      cur_data(),
      ~condition
    ) %>%
      DESeq(parallel = TRUE, BPPARAM = BiocParallel::MulticoreParam(workers = 6)) %>%
      list()
  )

de_controls <- de_meta %>%
  filter(agent == "DMSO") %>%
  distinct(
    source_96, cell_line, timepoint, control_condition = condition
  )

de_comps <- de_meta %>%
  filter(agent != "DMSO") %>%
  distinct(
    source_96, cell_line, agent, concentration, timepoint, condition
  ) %>%
  inner_join(
    de_controls
  )

de_results <- de_groups %>%
  inner_join(
    de_comps
  ) %>%
  rowwise() %>%
  mutate(
    de_res = results(
      des,
      contrast = c("condition", condition, control_condition),
      tidy = TRUE,
      parallel = TRUE, BPPARAM = BiocParallel::MulticoreParam(workers = 6)
    ) %>%
      dplyr::rename(ensembl_gene_id = row) %>%
      list()
  ) %>%
  ungroup()

```


