---
title: "CDK7 pathway decomposition"
author: "Clemens Hug"
date: "2025-06-17"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)
library(qs)
library(synExtra)
library(powerjoin)
library(decoupleR)
library(here)

synapser::synLogin()
syn <- synDownloader("~/data", .cache = TRUE)

syn_analysis <- "syn51221227"

plot_dir <- here("plots", "pathway_decomposition")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
```


```{r}
ensembl_gtf <- rtracklayer::import("Homo_sapiens.GRCh38.109.gtf.gz") %>%
  as_tibble()

deep_baseline_counts_long <- syn("syn60530854") %>%
  read_csv() %>%
  mutate(
    cell_line = str_replace_all(cell_line, fixed("-"), "")
  )

gr_raw <- syn("syn52242760") %>%
  read_csv() %>%
  mutate(neglog_GR50 = -log10(GR50))

deseq_dge_baseline_res <- syn("syn60527359") %>%
  read_csv() %>%
  mutate(
    across(cell_line, \(x) str_replace_all(x, fixed("-"), ""))
  )
```

```{r}
gr_long <- gr_raw %>%
  mutate(
    cell_line = str_replace_all(cell_line, fixed("-"), "")
  ) %>%
  select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
  pivot_longer(
    cols = -cell_line,
    names_to = "gr_metric",
    values_to = "gr_value"
  )

gr_mats <- gr_raw %>%
  mutate(
    cell_line = str_replace_all(cell_line, fixed("-"), "")
  ) %>%
  select(cell_line, GR50, GRmax, GR_AOC, neglog_GR50) %>%
  column_to_rownames("cell_line") %>%
  as.matrix() %>%
  t()
```

deep_baseline_counts_long %>% distinct(cell_line) %>% arrange(cell_line) %>% pull(cell_line)
c("59M", "CAOV-3", "COV318", "COV504", "ES-2", "EVSA-T", "FCI-P2p",
"FNE1", "FNE2", "FUOV1", "JHOM-1", "JHOS-2", "JHOS-4", "Kuramochi",
"OAW-28", "OCE1", "OCE2", "OCI-P2a", "OCI-P5x", "OCI-P8p", "OCI-U1a",
"ONCODG1", "OV-90", "OVCA-R4", "OVKATE", "OVSAHO", "RMUGS", "SNU-119",
"SNU-8", "TYK-NU")

gr_long %>% distinct(cell_line) %>% arrange(cell_line) %>% pull(cell_line) %>% dput()
c("59M", "CAOV3", "COV318", "COV362", "COV504", "ES2", "FCIP2p",
"FNE1", "FUOV1", "JHOM1", "JHOS2", "Kuramochi", "OAW28", "OCE1",
"OCE2", "OCIP5x", "OCIU1a", "ONCODG1", "OV90", "OVCAR3", "OVCAR4",
"OVCAR8", "OVKATE", "OVSAHO", "RMUGS", "SKOV3", "SNU119", "SNU8",
"TYKNU")


```{r}
intersect(
  deep_baseline_counts_long$cell_line,
  colnames(gr_mats)
)
setdiff(
  colnames(gr_mats),
  deep_baseline_counts_long$cell_line
)

```

Only 4 cell lines are missing in the counts data, but present in the GR data
"COV362" "OVCAR3" "OVCAR8" "SKOV3"

```{r}
deep_baseline_counts_long %>%
  distinct(ensembl_gene_id, hgnc_symbol) %>%
  count(hgnc_symbol) %>%
  count(n)

deep_baseline_counts_long %>%
  distinct(ensembl_gene_id, hgnc_symbol) %>%
  group_by(hgnc_symbol) %>%
  filter(n() > 1) %>%
  arrange(hgnc_symbol)

varstab_mat <- deep_baseline_counts_long %>%
  filter(
    type == "varstab",
    hgnc_symbol %in% {
      deep_baseline_counts_long %>%
        distinct(ensembl_gene_id, hgnc_symbol) %>%
        group_by(hgnc_symbol) %>%
        filter(n() == 1) %>%
        pull(hgnc_symbol)
    }
  ) %>%
  select(hgnc_symbol, cell_line, count) %>%
  pivot_wider(names_from = cell_line, values_from = count) %>%
  column_to_rownames("hgnc_symbol") %>%
  as.matrix()
```


```{r}
library(progeny)

progeny_res_perm <- progeny(
  varstab_mat,
  organism = "Human",
  scale = TRUE,
  verbose = TRUE,
  perm = 1000,
  get_nulldist = TRUE,
  z_scores = TRUE # Needs to be true, otherwise some useless normalized value is returned
)

progeny_res_perm_long <- progeny_res_perm[[1]] %>%
  magrittr::set_colnames(colnames(varstab_mat)) %>%
  as_tibble(rownames = "pathway") %>%
  pivot_longer(-pathway, names_to = "cell_line", values_to = "activity") %>%
  mutate(
    pathway = recode(pathway, `JAK.STAT` = "JAK-STAT")
  )

progeny_res_norm <- progeny(
  varstab_mat,
  organism = "Human",
  scale = FALSE,
  verbose = TRUE
)

progeny_res_norm_long <- progeny_res_norm %>%
  as_tibble(rownames = "cell_line") %>%
  pivot_longer(-cell_line, names_to = "pathway", values_to = "activity")
```

Perm version seems preferable, represents the pathway activity relative to 1000
random gene sets. The normal version is some sort of coefficient from a model.

```{r}
decoup_progeny_net <- get_progeny(organism = "human", top = 100)


decoup_progeny_ulm_res <- run_ulm(
  varstab_mat,
  decoup_progeny_net
)
```


```{r}
power_inner_join(
  progeny_res_norm_long,
  decoup_progeny_ulm_res,
  by = c("cell_line" = "condition", "pathway" = "source"),
  check = check_specs(
    unmatched_keys_left = "warn",
    unmatched_keys_right = "warn",
    duplicate_keys_left = "warn",
    duplicate_keys_right = "warn"
  )
) %>%
  ggplot(
    aes(
      activity, score
    )
  ) +
  geom_point() +
  facet_wrap(~cell_line)


power_inner_join(
  progeny_res_perm_long,
  decoup_progeny_ulm_res,
    # mutate(
    #   across(c(source, condition), \(x) str_replace_all(x, "[^A-Za-z0-9]", ".")),
    # ),
  by = c("cell_line" = "condition", "pathway" = "source"),
  check = check_specs(
    unmatched_keys_left = "warn",
    unmatched_keys_right = "warn",
    duplicate_keys_left = "warn",
    duplicate_keys_right = "warn"
  )
) %>%
  ggplot(
    aes(
      activity, score
    )
  ) +
  geom_point() +
  facet_wrap(~cell_line)


power_inner_join(
  progeny_res_perm_long,
  decoup_progeny_ulm_res,
    # mutate(
    #   across(c(source, condition), \(x) str_replace_all(x, "[^A-Za-z0-9]", ".")),
    # ),
  by = c("cell_line" = "condition", "pathway" = "source"),
  check = check_specs(
    unmatched_keys_left = "warn",
    unmatched_keys_right = "warn",
    duplicate_keys_left = "warn",
    duplicate_keys_right = "warn"
  )
) %>%
  ggplot(
    aes(
      activity, score
    )
  ) +
  geom_point() +
  facet_wrap(~pathway)
```


```{r}
library(ComplexHeatmap)

progeny_res_perm_hm <- progeny_res_perm_long %>%
  pivot_wider(names_from = cell_line, values_from = activity) %>%
  column_to_rownames("pathway") %>%
  as.matrix()

hm <- Heatmap(
  progeny_res_perm_hm,
  col = circlize::colorRamp2(
    {
      q1 <- quantile(progeny_res_perm_hm, 0.01, na.rm = TRUE)
      q99 <- quantile(progeny_res_perm_hm, 0.99, na.rm = TRUE)
      absmax <- max(abs(q1), abs(q99))
      seq(-absmax, absmax, length.out = 11)
    },
    rev(RColorBrewer::brewer.pal(11, "RdBu"))
  ),
  name = "Activity",
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
)
hm

withr::with_pdf(
  file.path(plot_dir, "progeny_perm_heatmap.pdf"),
  {
    draw(hm)
  }, width = 10, height = 6
)
 ```

## progeny vs GR values

```{r}
gr_vs_progeny_data <- power_inner_join(
  progeny_res_perm_long,
  gr_long,
  by = c("cell_line"),
  check = check_specs(
    unmatched_keys_left = "info",
    unmatched_keys_right = "info"
  )
)

```


```{r}
gr_vs_progeny_lms <- gr_vs_progeny_data %>%
  group_by(gr_metric, pathway) %>%
  summarize(
    lm_res = {
      lm(
        gr_value ~ activity,
        data = cur_data()
      ) %>%
        broom::tidy() %>%
        list()
    },
    .groups = "drop"
  ) %>%
  unnest(lm_res) %>%
  filter(term == "activity") %>%
  mutate(
    padj = p.adjust(p.value, method = "fdr")
  )



p <- ggplot(gr_vs_progeny_data, aes(x = activity, y = gr_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  ggh4x::facet_grid2(
    rows = vars(gr_metric),
    cols = vars(pathway),
    scales = "free",
    independent = "all"
  ) +
  theme_bw()
p
```


## glmnet

```{r}
gr_vs_progeny_glmnet_raw <- gr_vs_progeny_data %>%
  group_by(gr_metric) %>%
  summarize(
    res = {
      mat <- pivot_wider(
        cur_data(),
        names_from = pathway,
        values_from = activity
      ) %>%
        column_to_rownames("cell_line") %>%
        as.matrix()
      # browser()
      glmnet::cv.glmnet(
        x = mat[, colnames(mat) != "gr_value"],
        y = mat[, "gr_value"],
        alpha = 0.5, # Elastic net mixing parameter
        nfolds = 10,
        standardize = TRUE,
        intercept = TRUE,
        family = "gaussian"
      ) %>%
        list()
    },
    .groups = "drop"
  )
```

## fgsea DGE baseline

```{r}
library(fgsea)
library(msigdbr)

all_gene_sets <- msigdbr(species = "Homo sapiens")
selected_gene_sets <- all_gene_sets %>%
  filter(
    gs_collection == "H" |
      gs_subcollection == "CP:REACTOME"
  )

selected_gene_sets_fgsea <- selected_gene_sets %>%
  group_nest(gs_collection, gs_subcollection, gs_name) %>%
  mutate(
    gs = map(
      data,
      \(x) unique(x$ensembl_gene)
    ) %>%
      set_names(gs_name)
  )

selected_des <- deseq_dge_baseline_res %>%
  mutate(
    signed_p = -sign(log2FoldChange_MLE) * log10(pvalue)
  ) %>%
  filter(is.finite(signed_p), is.finite(log2FoldChange)) %>%
  group_nest(cell_line) %>%
  mutate(
    de = map(
      data,
      \(x) set_names(x$log2FoldChange, x$ensembl_gene_id)
    ),
    de_p = map(
      data,
      \(x) set_names(x$signed_p, x$ensembl_gene_id)
    )
  )

fgsea_res_raw <- selected_des %>%
  mutate(
    res = map(
      de,
      \(x) fgseaMultilevel(
        selected_gene_sets_fgsea$gs,
        x
      )
    ),
    res_p = map(
      de_p,
      \(x) fgseaMultilevel(
        selected_gene_sets_fgsea$gs,
        x
      )
    )
  )

fgsea_res_long <- fgsea_res_raw %>%
  select(cell_line, res) %>%
  unnest(res) %>%
  mutate(
    signed_p = -sign(NES) * log10(padj),
    p_cut = cut(padj, breaks = c(-Inf, 0.01, 0.05, Inf), labels = c("**", "*", ""))
  ) %>%
  filter(
    pathway %in% {
      group_by(., pathway) %>%
        summarize(n_sig = sum(padj < 0.05)) %>%
        filter(n_sig > 0) %>%
        pull(pathway)
    }
  ) %>%
  mutate(
    size_leading_edge = map_int(
      leadingEdge, length
    )
  ) %>%
  select(
    cell_line, pathway,
    pval, padj, p_cut, ES, NES, size_leading_edge, size,
    leadingEdge
  )

fgsea_res_vs_gr <- power_inner_join(
  fgsea_res_long,
  gr_long,
  by = c("cell_line"),
  check = check_specs(
    unmatched_keys_left = "info",
    unmatched_keys_right = "info"
  )
)
```

## glmnet

```{r}
gr_vs_fgsea_glmnet_raw <- fgsea_res_vs_gr %>%
  group_by(gr_metric) %>%
  summarize(
    res = {
      mat <- pivot_wider(
        pick(cell_line, pathway, NES, gr_value),
        names_from = pathway,
        values_from = NES
      ) %>%
        column_to_rownames("cell_line") %>%
        as.matrix()
      # browser()
      glmnet::cv.glmnet(
        x = mat[, colnames(mat) != "gr_value"],
        y = mat[, "gr_value"],
        alpha = 0.5, # Elastic net mixing parameter
        nfolds = 10,
        standardize = TRUE,
        intercept = TRUE,
        family = "gaussian"
      ) %>%
        list()
    },
    .groups = "drop"
  )

```


## GSVA

```{r}
varstab_mat_ensembl <- deep_baseline_counts_long %>%
  filter(
    type == "varstab"
  ) %>%
  select(ensembl_gene_id, cell_line, count) %>%
  pivot_wider(names_from = cell_line, values_from = count) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()

library(GSVA)

gsva_param <- gsvaParam(
  varstab_mat_ensembl,
  geneSets = selected_gene_sets_fgsea$gs
)

gsva_mat <- gsva(
  gsva_param
)

gsva_res_long <- gsva_mat %>%
  as_tibble(rownames = "pathway") %>%
  pivot_longer(-pathway, names_to = "cell_line", values_to = "activity")

gsva_res_vs_gr <- power_inner_join(
  gsva_res_long,
  gr_long,
  by = c("cell_line"),
  check = check_specs(
    unmatched_keys_left = "info",
    unmatched_keys_right = "info"
  )
)
```


## glmnet

```{r}
gr_vs_gsva_glmnet_raw <- gsva_res_vs_gr %>%
  group_by(gr_metric) %>%
  summarize(
    res = {
      mat <- pivot_wider(
        pick(cell_line, pathway, activity, gr_value),
        names_from = pathway,
        values_from = activity
      ) %>%
        column_to_rownames("cell_line") %>%
        as.matrix()
      # browser()
      glmnet::cv.glmnet(
        x = mat[, colnames(mat) != "gr_value"],
        y = mat[, "gr_value"],
        alpha = 0.5, # Elastic net mixing parameter
        nfolds = 10,
        standardize = TRUE,
        intercept = TRUE,
        family = "gaussian"
      ) %>%
        list()
    },
    .groups = "drop"
  )

gr_vs_gsva_glmnet_coefs <- gr_vs_gsva_glmnet_raw %>%
  mutate(
    coefs = map(
      res,
      \(x) {
        coef(x, s = "lambda.min") %>%
          as.matrix() %>%
          as_tibble(rownames = "pathway") %>%
          filter(pathway != "(Intercept)", s0 > 0)
      }
    )
  )

p <- ggplot(
  gsva_res_vs_gr %>%
    filter(gr_metric == "neglog_GR50") %>%
    power_inner_join(
      gr_vs_gsva_glmnet_coefs$coefs[[4]],
      by = "pathway",
      check = check_specs(
        unmatched_keys_right = "warn"
      )
    ),
  aes(
    x = activity, y = gr_value
  )
) +
  geom_point() +
  facet_wrap(~pathway, scales = "free")
p
```

