---
title: "CDK7 PRISM analysis"
author: "Clemens Hug"
date: "2026-02-06"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(powerjoin)
library(synExtra)
library(conflicted)
library(hdf5r)
library(ggbeeswarm)
library(here)
library(lmerTest)
library(emmeans)
library(broom.mixed)
library(furrr)
library(qs2)
library(ggrepel)
library(msigdbr)
library(fgsea)
library(openxlsx)
library(ggrastr)
library(ashr)

conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  base::setdiff,
  lmerTest::lmer
)

synapser::synLogin()
syn <- synDownloader("~/data", .cache = TRUE)

plot_dir <- here("plots", "prism")
dir.create(plot_dir, showWarnings = FALSE)

result_dir <- here("prism")
dir.create(result_dir, showWarnings = FALSE)

set_theme(
  envalysis::theme_publish(
    base_family = "Helvetica",
    base_size = 10
  )
)
```

PRISM analysis code at https://github.com/theprismlab/sushi/blob/5320078b788fd606ec562ac60ebdb4c9688087f5/scripts/biomarker/biomarker_functions.R#L103

## Load Biomarker data

```{r}
biomarkers_raw <- syn("syn73615969") |>
  H5File$new(mode = "r")

expression_col_meta <- biomarkers_raw[["Expression/column_meta"]]$read() |>
  as_tibble()

expression_row_meta <- biomarkers_raw[["Expression/row_meta"]]$read() |>
  as_tibble()

expression_mat <- biomarkers_raw[["Expression/mat"]]$read() |>
  magrittr::set_colnames(
    expression_col_meta$column_name
  ) |>
  magrittr::set_rownames(
    expression_row_meta$ModelID
  )

cell_line_meta <- biomarkers_raw[["Lineage_Table/table"]]$read() |>
  as_tibble()

# for mapping old YKL ccle names to new depmap ids
cell_line_meta_depmap <- syn("syn73674394") |>
  read_csv()
```


```{r}
prism_raw_zips <- c(
  mts_30 = "syn73616950",
  ykl = "syn73616961"
) |>
  map(syn)

prism_drc <- list(
  mts_30 = prism_raw_zips[["mts_30"]] |>
    unz("drc/MTS030_TINGHU_ZHANG_DRC_TABLE.csv") |>
    read_csv(),
  ykl = prism_raw_zips[["ykl"]] |>
    unz("data/DRC_TABLE.csv") |>
    read_csv() |>
    rename(
      minimum_dose = min_dose,
      maximum_dose = max_dose,
      inflection = ec50,
      log2_ic50 = log2.ic50,
      frac_var_explained = R2
    ) |>
    power_inner_join(
      distinct(cell_line_meta_depmap, ccle_name = CCLEName, depmap_id = ModelID) |>
        drop_na(),
      by = "ccle_name",
      check = check_specs(
        unmatched_keys_left = "warn",
        duplicate_keys_right = "warn"
      )
    )
) |>
  bind_rows(
    .id = "dataset"
  )


```


## Explore MDR genes

```{r}
mdr_genes <- c("ABCB1", "ABCG2", "ABCC1")

mdr_expression <- expression_mat[, mdr_genes] |>
  as_tibble(rownames = "depmap_id") |>
  pivot_longer(
    -depmap_id,
    names_to = "gene",
    values_to = "expression"
  ) |>
  power_left_join(
    cell_line_meta,
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  )

p <- ggplot(
  mdr_expression |>
    mutate(
      across(lineage, fct_rev)
    ),
  aes(
    x = expression,
    y = lineage,
    color = lineage
  )
) +
  geom_vline(
    aes(
      xintercept = xintercept
    ),
    data = tibble(gene = "ABCB1", xintercept = 1),
    linetype = "dashed",
    color = "black"
  ) +
  geom_quasirandom(
    shape = 16
  ) +
  facet_wrap(
    ~gene,
    scales = "free_x",
    axes = "all", axis.labels = "margins"
  ) +
  guides(color = "none") +
  scale_x_continuous(
    expand = expansion(mult = c(0.01, .05))
  ) +
  labs(
    x = "Expression", y = NULL
  )
p

ggsave(
  file.path(plot_dir, "mdr_genes_expression.pdf"),
  p, width = 10, height = 6.5
)
```

Check correlation of the three

A sensible cutoff for ABCB1 seems to be around 1-2. Clear negative population with
long positive tail. For ABCG2 bit harder to define. Much fatter tail especially
for lung and skin. ABCC1 doesn't have negative population at all.

Just doing ABCB1 cutoff 1 for now.

## Filter cell lines and genes

```{r}
ABCB1_CUTOFF <- 1

cell_lines_n_profiled <- prism_drc |>
  distinct(depmap_id, pert_name) |>
  count(depmap_id)

cell_lines_n_profiled |>
  count(n)

cell_lines_filtered <- cell_line_meta |>
  semi_join(
    mdr_expression |>
      filter(gene == "ABCB1", expression <= ABCB1_CUTOFF),
    by = "depmap_id"
  ) |>
  semi_join(
    cell_lines_n_profiled |>
      filter(n >= 2),
    by = "depmap_id"
  )


```

Selecting genes with sufficient expression in at least some cell lines.

```{r}
eligible_genes <- apply(
  expression_mat, 2,
  \(x) quantile(x, .8, names = FALSE)
) > 1

```

## Compound correlation

How well are AUC values correlated between the three compounds?

```{r}
auc_comparison <- combn(
  unique(prism_drc$pert_name), 2
) |>
  t() |>
  magrittr::set_colnames(c("compound_1", "compound_2")) |>
  as_tibble() |>
  inner_join(
    select(prism_drc, depmap_id, pert_name, auc),
    by = c("compound_1" = "pert_name"),
    relationship = "many-to-many"
  ) |>
  inner_join(
    select(prism_drc, depmap_id, pert_name, auc),
    by = c("compound_2" = "pert_name", "depmap_id"),
    suffix = c("_1", "_2"),
    relationship = "many-to-many"
  ) |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  mutate(
    filtered = if_else(
      depmap_id %in% cell_lines_filtered$depmap_id,
      "Included",
      "Excluded"
    )
  )

p <- ggplot(
  auc_comparison,
  aes(
    x = 1 - auc_1,
    y = 1 - auc_2,
    color = lineage
  )
) +
  geom_abline(
    slope = 1, intercept = 0,
    linetype = "dashed",
    color = "grey"
  ) +
  geom_point(
    aes(
      alpha = filtered
    ),
    shape = 16
  ) +
  geom_text(
    aes(
      label = r
    ),
    data = \(x) group_by(x, compound_1, compound_2) |>
      drop_na(auc_1, auc_2) |>
      summarize(
        r = paste0(
          "  R = ", signif(cor(auc_1, auc_2), 2), " (all data)\n",
          "  R = ", signif(cor(auc_1[filtered == "Included"], auc_2[filtered == "Included"]), 2), " (ABCB1 filtered)"
        ),
        .groups = "drop"
      ),
    color = "black",
    x = -Inf, y = Inf,
    hjust = 0, vjust = 1
  ) +
  scale_color_discrete(
    palette = \(n) scales::alpha(scales::hue_pal()(n), .6),
    guide = "none"
  ) +
  scale_alpha_manual(
    values = c(
      "Included" = .9,
      "Excluded" = .2
    ),
    name = "ABCB1 filtering"
  ) +
  facet_wrap(
    ~paste(compound_1, "vs", compound_2),
    axes = "all", axis.labels = "margins"
  ) +
  labs(
    x = "AOC Compound 1", y = "AOC Compound 2"
  ) +
  coord_equal()
p

ggsave(
  file.path(plot_dir, "auc_comparison.pdf"),
  p, width = 7, height = 4
)
```

## Fitting lineage model

Removing lineages that don't have data for all three compounds.
Otherwise the model won't be able to fit the interaction term properly.
Only affects few cells lines, Adrenal gland, Eye, and Vulva/Vagina.

```{r}
prism_drc_lineage <- prism_drc |>
  select(depmap_id, pert_name, auc) |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  group_by(lineage) |>
  filter(
    n_distinct(pert_name) == 3
  ) |>
  ungroup()

lineage_vector <- cell_line_meta |>
  semi_join(
    prism_drc_lineage,
    by = "depmap_id"
  ) |>
  group_nest(lineage) |>
  mutate(
    gs = set_names(
      map(data, "depmap_id"),
      lineage
    )
  )

prism_fgsea_res <- prism_drc_lineage |>
  drop_na(auc) |>
  group_by(pert_name, depmap_id) |>
  summarize(
    auc = mean(auc[is.finite(auc)], na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(pert_name) |>
  summarize(
    res = fgseaMultilevel(
      lineage_vector$gs,
      set_names(scale(auc)[, 1], depmap_id)
    ) |>
      arrange(padj, pval) |>
      list()
  )
```

```{r}
prism_lineage_model <- lm(
  auc ~ lineage * pert_name,
  data = prism_drc_lineage |>
    filter(depmap_id %in% cell_lines_filtered$depmap_id)
)
car::Anova(prism_lineage_model)

lineage_emm <- emmeans(
  prism_lineage_model,
  ~ lineage | pert_name
)

lineage_emm_res <- contrast(lineage_emm, method = "eff", adjust = "fdr") |>
  as_tibble() |>
  arrange(pert_name, p.value)
```


```{r}
prism_lineage_model_sans_filtering <- lm(
  auc ~ lineage * pert_name,
  data = prism_drc_lineage
)
car::Anova(prism_lineage_model_sans_filtering)

lineage_emm_sans_filtering <- emmeans(
  prism_lineage_model_sans_filtering,
  ~ lineage | pert_name
)

lineage_emm_res_sans_filtering <- contrast(lineage_emm_sans_filtering, method = "eff", adjust = "fdr") |>
  as_tibble() |>
  arrange(pert_name, p.value)
```


```{r}
p <- ggplot(
  prism_drc_lineage |>
    mutate(
      across(lineage, fct_rev),
      filtered = if_else(
        depmap_id %in% cell_lines_filtered$depmap_id,
        "Included",
        "Excluded"
      )
    ),
  aes(
    x = 1 - auc,
    y = lineage,
    color = lineage
  )
) +
  geom_quasirandom(
    aes(
      color = lineage,
      alpha = filtered
    ),
    shape = 16
  ) +
  scale_alpha_manual(
    values = c(
      "Included" = .9,
      "Excluded" = .2
    ),
    name = "MDR filtering"
  ) +
  # geom_text(
  #   aes(
  #     label = paste0(
  #       if_else(estimate > 0, "  ", ""),
  #       cut(p.value, c(-Inf, 0.001, 0.01, 0.05, Inf), c("***", "**", "*", "")),
  #       if_else(estimate > 0, "", "  ")
  #     ),
  #     x = if_else(estimate > 0, -Inf, Inf),
  #     hjust = if_else(estimate > 0, 0, 1),
  #   ),
  #   data = lineage_emm_res_sans_filtering |>
  #     mutate(
  #       lineage = str_remove(contrast, " effect") |>
  #         str_remove_all("\\(|\\)")
  #     ),
  #   color = "black",
  #   vjust = .5
  # ) +
  scale_color_discrete(
    palette = \(n) scales::alpha(scales::hue_pal()(n), .6),
    guide = "none"
  ) +
  geom_boxplot(
    data = \(df) filter(df, depmap_id %in% cell_lines_filtered$depmap_id),
    color = "black",
    outlier.shape = NA,
    fill = NA
  ) +
  facet_wrap(
    ~pert_name,
    scales = "free_x",
    axes = "all", axis.labels = "margins"
  ) +
  labs(
    x = "Area over curve (AOC)", y = NULL,
    color = ""
  )
p

ggsave(
  file.path(plot_dir, "auc_by_lineage_alpha.pdf"),
  p, width = 8, height = 4.5
)
```

## Fitting expression model

The goal is to fit a linear model of AUC ~ expression * drug

```{r}
expression_mat_ifit_composite <- cbind(
  expression_mat,
  expression_mat[, c("IFITM1", "IFITM2", "IFITM3")] |>
    scale() |>
    apply(1, mean) |>
    as.matrix(ncol = 1) |>
    magrittr::set_colnames("IFITM-composite")
)

input_data <- prism_drc |>
  filter(
    depmap_id %in% cell_lines_filtered$depmap_id
  ) |>
  select(depmap_id, pert_name, auc)  |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  power_left_join(
    expression_mat_ifit_composite[
      cell_lines_filtered$depmap_id,
      c(eligible_genes, c("IFITM-composite" = TRUE))
    ] |>
      as_tibble(rownames = "depmap_id") |>
      pivot_longer(
        -depmap_id,
        names_to = "gene",
        values_to = "expression"
      ),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn"
    )
  ) |>
  group_nest(gene)
```

```{r}
run_model <- function(
  df, model_formula, random_effects = TRUE,
  lmer.df = "satterthwaite", REML = FALSE,
  rank_transform = FALSE
) {
  # browser()
  if (rank_transform) {
    df <- df |>
      mutate(
        across(
          c(auc, expression),
          rank
        )
      )
  }
  if (random_effects) {
    model <- lmer(
      formula(model_formula),
      data = df,
      REML = REML
    )
  } else {
    model <- lm(
      formula(model_formula),
      data = df
    )
  }

  emm <- emtrends(
    model,
    ~pert_name,
    var = "expression",
    adjust = "none",
    lmer.df = lmer.df
  )

  emm_intercepts <- emmeans(
    model,
    ~pert_name,
    at = list(expression = 0),
    adjust = "none",
    lmer.df = lmer.df
  )

  res_by_compound <- inner_join(
    tidy(emm_intercepts) |>
      rename(intercept = estimate) |>
      select(-c(df, statistic, p.value)),
    tidy(emm),
    by = "pert_name",
    suffix = c("_intercept", "_slope")
  )

  list(
    model_stats = glance(model),
    model_coefs = tidy(model),
    model_by_compound = res_by_compound,
    pairwise_res = pairs(
      emm,
      adjust = "none"
    ) |>
      tidy()
  )
}

```

```{r}
model_formulas <- tribble(
  ~model_name, ~model_formula, ~random_effects,
  "random_lineage", auc ~ expression * pert_name + (1 | lineage), TRUE,
  "random_cell_line", auc ~ expression * pert_name + (1 | depmap_id), TRUE,
  "fixed", auc ~ expression * pert_name, FALSE
)

test_output <- input_data |>
  slice(1) |>
  crossing(model_formulas) |>
  mutate(
    model_out = pmap(
      list(data, model_formula, random_effects),
      \(d, f, r) run_model(
        d,
        f,
        r,
        REML = FALSE
      )
    )
  )

plan(multicore, workers = 4)

model_res_raw_all <- input_data |>
  crossing(model_formulas) |>
  mutate(
    model_out = future_pmap(
      list(data, model_formula, random_effects),
      \(data, model_formula, random_effects) run_model(
        data,
        model_formula,
        random_effects,
        REML = FALSE
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )

qs_save(
  model_res_raw_all,
  file.path(result_dir, "model_res_all_no_reml.qs")
)
# model_res_raw <- qd_read(
#   file.path(result_dir, "model_res_raw.qs")
# )


model_res_aic <- model_res_raw_all |>
  transmute(
    gene, model_name,
    aic = map_dbl(
      model_out, \(x) x$model_stats$AIC
    ),
    bic = map_dbl(
      model_out, \(x) x$model_stats$BIC
    )
  ) |>
  group_by(gene) |>
  mutate(
    rank_aic = rank(-aic),
    rank_bic = rank(-bic)
  ) |>
  ungroup()

model_res_aic_summary <- model_res_aic |>
  group_by(model_name) |>
  summarize(
    n_best_aic = sum(rank_aic == 1),
    n_best_bic = sum(rank_bic == 1),
    mean_aic = mean(rank_aic),
    mean_bic = mean(rank_bic),
    .groups = "drop"
  )

```

  model_name       n_best_aic n_best_bic mean_aic mean_bic
  <chr>                 <int>      <int>    <dbl>    <dbl>
1 fixed                  3852         44     1.71     2.00
2 random_cell_line          0          0     3        3
3 random_lineage         9424      13232     1.29     1.00

Running the random_lineage model

```{r}
plan(multicore, workers = 8)

model_res_raw <- input_data |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger"
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )

qs_save(
  model_res_raw,
  file.path(result_dir, "model_res_raw.qs")
)
# model_res_raw <- qs_read(
#   file.path(result_dir, "model_res_raw.qs")
# )

model_res_rank_raw <- input_data |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger",
        rank_transform = TRUE
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )

qs_save(
  model_res_rank_raw,
  file.path(result_dir, "model_res_rank_raw.qs")
)
```


```{r}
model_res_pairwise_shrunk <- model_res_raw |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  group_nest(
    term, contrast
  ) |>
  mutate(
    shrunk_pairwise_res = map(
      data,
      \(df) ash(
        df$estimate, df$std.error,
        df = NULL,
        mixcompdist = "uniform"
      )
    ),
    data = map2(
      data, shrunk_pairwise_res,
      \(df, ash_res) {
        df |>
          mutate(
            estimate_shrunk = ash_res$result$PosteriorMean,
            std.error_shrunk = ash_res$result$PosteriorSD,
            s_value_shrunk = ash_res$result$svalue,
            q_value_shrunk = ash_res$result$qvalue,
            padj = p.adjust(p.value, method = "fdr")
          )
      }
    )
  )

model_res_pairwise <- model_res_pairwise_shrunk |>
  transmute(
    contrast,
    data
  ) |>
  unnest(data) |>
  mutate(
    direction = if_else(estimate > 0, "positive", "negative")
  ) |>
  separate_wider_delim(
    contrast, delim = " - ",
    names = c("pert_name_1", "pert_name_2"),
    cols_remove = FALSE
  ) |>
  mutate(
    across(
      starts_with("pert_name"),
      \(x) str_remove_all(x, "\\(|\\)")
    )
  )


model_slopes_shrunk <- model_res_raw |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  group_nest(pert_name) |>
  mutate(
    shrunk_model_slopes = map(
      data,
      \(df) ash(
        df$expression.trend, df$std.error_slope,
        df = NULL,
        mixcompdist = "uniform"
      )
    ),
    data = map2(
      data, shrunk_model_slopes,
      \(df, ash_res) {
        df |>
          mutate(
            expression.trend_shrunk = ash_res$result$PosteriorMean,
            std.error_shrunk = ash_res$result$PosteriorSD,
            s_value_shrunk = ash_res$result$svalue,
            q_value_shrunk = ash_res$result$qvalue,
            padj = p.adjust(p.value, method = "fdr")
          )
      }
    )
  )

model_slopes <- model_slopes_shrunk |>
  transmute(
    pert_name,
    data
  ) |>
  unnest(data) |>
  mutate(
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```

```{r}
model_res_pairwise_rank <- model_res_rank_raw |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(estimate > 0, "positive", "negative")
  )

model_slopes_rank <- model_res_rank_raw |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```

### Without MDR filtering

```{r}
cell_lines_sans_mdr_filtering <- cell_line_meta |>
  filter(
    depmap_id %in% rownames(expression_mat)
  ) |>
  semi_join(
    cell_lines_n_profiled |>
      filter(n >= 2),
    by = "depmap_id"
  )

input_data_sans_mdr_filtering <- prism_drc |>
  filter(
    depmap_id %in% cell_lines_sans_mdr_filtering$depmap_id
  ) |>
  select(depmap_id, pert_name, auc)  |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  power_left_join(
    expression_mat_ifit_composite[
      cell_lines_sans_mdr_filtering$depmap_id,
      c(eligible_genes, c("IFITM-composite" = TRUE))
    ] |>
      as_tibble(rownames = "depmap_id") |>
      pivot_longer(
        -depmap_id,
        names_to = "gene",
        values_to = "expression"
      ),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn"
    )
  ) |>
  group_nest(gene)

```


```{r}
plan(multicore, workers = 6)

model_res_raw2 <- input_data_sans_mdr_filtering |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger"
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )
qs_save(
  model_res_raw2,
  file.path(result_dir, "model_res_raw_no_mdr_filtering.qs")
)
# model_res_raw2 <- qs_read(
#   file.path(result_dir, "model_res_raw_no_mdr_filtering.qs")
# )

model_res_rank_raw <- input_data_sans_mdr_filtering |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger",
        rank_transform = TRUE
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )
qs_save(
  model_res_rank_raw,
  file.path(result_dir, "model_res_rank_raw_no_mdr_filtering.qs")
 )
```


* Fraction of cell lines that respond for each compound (AUC < .25)
* Overview of ABCB1 and IFITM expression
* ABCB1/AUC correlation
* Plot correlation IFIT/AUC
* Make composite IFITM score


```{r}
model_mdr_included_res_pairwise_ <- model_res_raw2 |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(estimate > 0, "positive", "negative")
  )

model_mdr_included_slopes <- model_res_raw2 |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```


## Plot

### P-value distribution

```{r}
p <- bind_rows(
  `Pairwise comparisons` = model_res_pairwise,
  `Model slopes` = model_slopes,
  .id = "result_type"
) |>
  ggplot(
    aes(x = p.value)
  ) +
  geom_histogram() +
  scale_x_continuous(
    expand = expansion(mult = c(0, .05))
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .05))
  ) +
  facet_wrap(
    ~result_type,
    scales = "free_y",
    axes = "all", axis.labels = "margins"
  ) +
  envalysis::theme_publish()

ggsave(
  file.path(plot_dir, "p_value_distributions.pdf"),
  p, width = 8, height = 4
)
```


```{r}
plot_auc_expression_corr <- function(gene, rank_transform = FALSE, mdr_filtering = TRUE) {
  if (mdr_filtering) {
    id <- input_data
  } else {
    id <- input_data_sans_mdr_filtering
  }
  scatter_df <- id |>
    filter(.data$gene == .env$gene) |>
    unnest(data) |>
    select(-lineage) |>
    power_inner_join(
      cell_line_meta,
      by = "depmap_id",
      check = check_specs(
        unmatched_keys_left = "warn",
        duplicate_keys_right = "warn"
      )
    )
  if (rank_transform) {
    scatter_df <- scatter_df |>
      mutate(
        across(
          c(auc, expression),
          rank
        )
      )
    slope_df <- model_slopes_rank |>
      filter(.data$gene == .env$gene)
  } else {
    if (mdr_filtering) {
      slope_df <- model_slopes |>
        filter(.data$gene == .env$gene)
    } else {
      slope_df <- model_mdr_included_slopes |>
        filter(.data$gene == .env$gene)
    }
  }
  ggplot(
    scatter_df,
    aes(
      x = expression,
      y = 1 - auc
    )
  ) +
    geom_point(
      aes(
        color = lineage
      ),
      alpha = .5,
      shape = 16
    ) +
    geom_abline(
      aes(
        slope = -expression.trend,
        intercept = 1 - intercept
      ),
      data = slope_df
    ) +
    ggnewscale::new_scale_color() +
    geom_text(
      aes(
        label = paste0(
          " b = ", signif(-expression.trend, 2),
          "\n p = ", signif(padj, 2)
        ),
        color = padj < 0.05
      ),
      data = slope_df,
      x = -Inf, y = Inf,
      hjust = 0, vjust = 1
    ) +
    scale_color_manual(
      values = c(
        "TRUE" = "red",
        "FALSE" = "black"
      ),
      guide = "none"
    ) +
    labs(
      x = "Expression", y = "AOC",
      color = "Lineage",
      title = gene
    ) +
    facet_wrap(
      ~pert_name,
      axes = "all", axis.labels = "margins"
    )
}

plot_auc_expression_corr("ABCB1", mdr_filtering = FALSE)
plot_auc_expression_corr("IFITM-composite", mdr_filtering = FALSE)
plot_auc_expression_corr("WDR54")
plot_auc_expression_corr("IFITM1")
plot_auc_expression_corr("AURKB")
plot_auc_expression_corr("CDK7")
plot_auc_expression_corr("MED12L")
plot_auc_expression_corr("PPIA")
plot_auc_expression_corr("IFIT1")
plot_auc_expression_corr("ZNF775")

plot_auc_expression_corr("PAQR3", rank_transform = TRUE)

input_data |>
  filter(gene == "ABCB1") |>
  chuck("data", 1L) |>
  plot_auc_expression_corr()

dir.create(
  file.path(plot_dir, "gene_aoc_correlation"),
  showWarnings = FALSE, recursive = TRUE
)

p <- plot_auc_expression_corr("ABCB1", mdr_filtering = FALSE)
ggsave(
  file.path(plot_dir, "gene_aoc_correlation", "ABCB1.pdf"),
  p, width = 7, height = 4
)

p <- plot_auc_expression_corr("IFITM-composite")
ggsave(
  file.path(plot_dir, "gene_aoc_correlation", "IFITM-composite.pdf"),
  p, width = 7, height = 4
)
```

Make plots for top 20 hits.


```{r}
dir.create(
  file.path(plot_dir, "top_hits", "pairwise"),
  showWarnings = FALSE, recursive = TRUE
)
dir.create(
  file.path(plot_dir, "top_hits", "slopes"),
  showWarnings = FALSE, recursive = TRUE
)

model_res_pairwise |>
  arrange(p.value) |>
  filter(padj < 0.05) |>
  group_by(contrast) |>
  slice_head(n = 20) |>
  mutate(
    rank = seq_len(n())
  ) |>
  ungroup() |>
  pwalk(
    \(gene, contrast, rank, ...) {
      p <- plot_auc_expression_corr(gene)
      ggsave(
        file.path(plot_dir, "top_hits", "pairwise", paste0(contrast, "_", rank, "_", gene, ".pdf")),
        p, width = 8, height = 6
      )
    }
  )

model_slopes |>
  arrange(p.value) |>
  filter(padj < 0.05) |>
  group_by(pert_name) |>
  slice_head(n = 20) |>
  mutate(
    rank = seq_len(n())
  ) |>
  ungroup() |>
  pwalk(
    \(gene, pert_name, rank, ...) {
      p <- plot_auc_expression_corr(gene)
      ggsave(
        file.path(plot_dir, "top_hits", "slopes", paste0(pert_name, "_", rank, "_", gene, ".pdf")),
        p, width = 8, height = 6
      )
    }
  )
```


## Volcano plot

```{r}
p <- ggplot(
  model_slopes |>
    mutate(
      p_censored = -log10(padj) > 12,
      neglog10_padj = pmin(-log10(padj), 12),
      signif_direction = case_when(
        padj > 0.05 ~ "not significant",
        expression.trend > 0 ~ "positive",
        expression.trend < 0 ~ "negative",
        TRUE ~ NA_character_
      )
    ) |>
    arrange(desc(padj), desc(p.value)),
  aes(
    x = expression.trend,
    y = neglog10_padj
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey"
  ) +
  rasterize(geom_point(
    aes(
      color = signif_direction,
      shape = p_censored
    ),
    alpha = .6
  ), dpi = 300, dev = "ragg_png") +
  geom_text_repel(
    data = \(x) arrange(x, p.value) |>
      filter(padj < 0.05) |>
      mutate(
        direction = if_else(
          expression.trend > 0, "positive", "negative"
        )
      ) |>
      group_by(pert_name, direction) |>
      slice_head(n = 20),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e6,
    seed = 42
  ) +
  scale_shape_manual(
    values = c(
      `TRUE` = 17,
      `FALSE` = 16
    ),
    guide = "none"
  ) +
  scale_color_manual(
    values = c(
      "positive" = paletteer::paletteer_c("ggthemes::Orange-Blue Diverging", 10)[3],
      "negative" = paletteer::paletteer_c("ggthemes::Orange-Blue Diverging", 10)[10],
      "not significant" = "black"
    ),
    guide = guide_legend(
      override.aes = list(
        shape = 16
      )
    )
    # values = c(
    #   "TRUE" = "red",
    #   "FALSE" = "black"
    # ),
    # guide = "none"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .05))
  ) +
  facet_wrap(
    ~pert_name,
    axes = "all", axis.labels = "margins",
    scales = "free_y"
  ) +
  labs(
    x = "Slope", y = "-log10(adj. p-value)"
  )
p

ggsave(
  file.path(plot_dir, "volcano_slopes.pdf"),
  p, width = 7, height = 4
)
```

```{r}
p <- ggplot(
  model_res_pairwise |>
    mutate(
      p_censored = -log10(padj) > 10,
      neglog10_padj = pmin(-log10(padj), 10),
      signif_direction = case_when(
        padj > 0.05 ~ "not significant",
        estimate > 0 ~ "positive",
        estimate < 0 ~ "negative",
        TRUE ~ NA_character_
      )
    ) |>
    arrange(desc(padj), desc(p.value)),
  aes(
    x = estimate,
    y = neglog10_padj
  )
) +
  rasterize(geom_point(
    aes(
      color = signif_direction,
      shape = p_censored
    ),
    alpha = .6
  ), dpi = 300, dev = "ragg_png") +
  geom_text_repel(
    data = \(x) arrange(x, p.value) |>
      mutate(
        direction = if_else(
          estimate > 0, "positive", "negative"
        )
      ) |>
      filter(padj < 0.05) |>
      group_by(contrast, direction) |>
      slice_head(n = 20),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e6,
    seed = 42
  ) +
  scale_shape_manual(
    values = c(
      `TRUE` = 17,
      `FALSE` = 16
    ),
    guide = "none"
  ) +
  scale_color_manual(
    values = c(
      "positive" = paletteer::paletteer_c("ggthemes::Orange-Blue Diverging", 10)[3],
      "negative" = paletteer::paletteer_c("ggthemes::Orange-Blue Diverging", 10)[10],
      "not significant" = "black"
    )
    # values = c(
    #   "TRUE" = "red",
    #   "FALSE" = "black"
    # ),
    # guide = "none"
  ) +
  facet_wrap(
    ~contrast,
    axes = "all", axis.labels = "margins",
    scales = "free_y"
  ) +
  labs(
    x = "Slope difference", y = "-log10(adj. p-value)"
  )
p

ggsave(
  file.path(plot_dir, "volcano_pairwise.pdf"),
  p, width = 7, height = 3.5
)
```


### Without MDR filtering

```{r}
p <- ggplot(
  model_mdr_included_slopes,
  aes(
    x = expression.trend,
    y = -log10(padj)
  )
) +
  rasterize(geom_point(
    aes(
      color = padj < 0.05
    ),
    shape = 16,
    alpha = .6
  ), dpi = 300, dev = "ragg_png") +
  geom_text_repel(
    data = \(x) arrange(x, p.value) |>
      filter(padj < 0.05) |>
      mutate(
        direction = if_else(
          expression.trend > 0, "positive", "negative"
        )
      ) |>
      group_by(pert_name, direction) |>
      slice_head(n = 20),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = "red",
      "FALSE" = "black"
    ),
    guide = "none"
  ) +
  facet_wrap(
    ~pert_name,
    axes = "all", axis.labels = "margins",
    scales = "free"
  ) +
  labs(
    x = "Slope", y = "-log10(adj. p-value)"
  )
p

ggsave(
  file.path(plot_dir, "volcano_slopes_no_filtering.pdf"),
  p, width = 7, height = 4
)
```


## Single compound vs comparison


```{r}
model_res_pairwise_with_rev <- model_res_pairwise |>
  bind_rows(
    transmute(
      model_res_pairwise,
      gene,
      pert_name_temp = pert_name_1,
      pert_name_1 = pert_name_2,
      pert_name_2 = pert_name_temp,
      contrast = paste0("(", pert_name_1, ") - (", pert_name_2, ")"),
      across(
        c(estimate, estimate_shrunk, statistic),
        \(x) -x
      ),
      null.value, df, std.error, p.value, s_value_shrunk, q_value_shrunk, padj,
      across(
        direction,
        \(x) if_else(x == "positive", "negative", "positive")
      )
    ) |>
      select(-pert_name_temp)
  )

res_pairs_vs_single <- power_inner_join(
    model_res_pairwise_with_rev,
    model_slopes,
    by = c("gene", "pert_name_1" = "pert_name"),
    check = check_specs(
      unmatched_keys_left = "warn",
      unmatched_keys_right = "warn",
      duplicate_keys_right = "warn"
    ),
    suffix = c("_pairwise", "_slope")
  ) |>
  mutate(
    direction_combined = case_when(
      padj_pairwise > .05 ~ "not significant",
      direction_pairwise == "positive" & direction_slope == "positive" ~ "both positive",
      direction_pairwise == "negative" & direction_slope == "negative" ~ "both negative",
      direction_pairwise == "positive" & direction_slope == "negative" ~ "positive negative",
      direction_pairwise == "negative" & direction_slope == "positive" ~ "negative positive",
      TRUE ~ NA_character_
    ) |>
      factor(
        levels = c(
          "both positive",
          "both negative",
          "positive negative",
          "negative positive",
          "not significant"
        )
      )
  )


p <- ggplot(
  res_pairs_vs_single |>
    filter(
      pert_name_1 %in% c("YNW-10-135", "YKL-5-124"),
      pert_name_2 %in% c("YNW-10-135", "YKL-5-124")
    ) |>
    arrange(desc(padj_pairwise), desc(p.value_pairwise)),
  aes(
    x = expression.trend_shrunk,
    y = estimate_shrunk
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  rasterize(geom_point(
    aes(
      color = direction_combined,
      text = gene
    ),
    alpha = .6,
    shape = 16
  ), dpi = 300, dev = "ragg_png") +
  geom_text_repel(
    # data = \(df) {
    #   # browser()
    #   arrange(df, padj_pairwise, p.value_pairwise) |>
    #   group_by(pert_name, direction_pairwise, direction_slope) |>
    #   mutate(
    #     gene_label = if_else(
    #       seq_len(n()) <= 10 & padj_pairwise < 0.05,
    #       gene, ""
    #     )
    #   ) |>
    #   ungroup()
    # },
    data = \(df) {
      arrange(df, padj_pairwise, p.value_pairwise) |>
        group_by(pert_name_1, pert_name_2, direction_combined) |>
        filter(padj_pairwise < 0.05) |>
        slice_head(n = 10) |>
        ungroup() |>
        mutate(
          gene_label = gene
        )
    },
    aes(label = gene_label),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e7,
    seed = 42,
    min.segment.length = 0
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(.1, .1))
  ) +
  scale_color_manual(
    values = c(
      "both positive" = "#e7967d",
      "both negative" =  "#2B5C8AFF",
      "positive negative" = "#e84271",
      "negative positive" = "#266B3EFF",
      "not significant" = "black"
    )
  ) +
  facet_wrap(
    ~paste(pert_name_1, "vs", pert_name_2),
    axes = "all", axis.labels = "margins",
    scales = "free"
  ) +
  labs(
    x = "Slope\nAOC vs gene expression",
    y = "Pairwise slope difference\n(YNW-10-135 vs YKL-5-124)"
  ) +
  theme(
    strip.clip = "off"
  )

ggsave(
  file.path(plot_dir, "pairwise_vs_slopes.pdf"),
  p, width = 4, height = 3
)

library(plotly)

ggplotly(p, tooltip = "text")
```

```{r}

res_pairs_vs_single_both <- power_inner_join(
    model_res_pairwise_with_rev,
    model_slopes,
    by = c("gene", "pert_name_1" = "pert_name"),
    check = check_specs(
      unmatched_keys_left = "warn",
      unmatched_keys_right = "warn",
      duplicate_keys_right = "warn"
    ),
    suffix = c("_pairwise", "")
) |>
  power_inner_join(
    model_slopes,
    by = c("gene", "pert_name_2" = "pert_name"),
    check = check_specs(
      unmatched_keys_left = "warn",
      unmatched_keys_right = "warn",
      duplicate_keys_right = "warn"
    ),
    suffix = c("_1", "_2")
  ) |>
  mutate(
    direction_pairwise_sign = case_when(
      padj_pairwise > .05 | (padj_1 > .05 & padj_2 > .05) ~ "not significant",
      direction_pairwise == "positive" ~ "positive",
      direction_pairwise == "negative" ~ "negative",
      TRUE ~ NA_character_
    )
  )

p <- ggplot(
  res_pairs_vs_single_both |>
    filter(
      pert_name_1 == "YNW-10-135",
      pert_name_2 == "YKL-5-124"
    ) |>
    arrange(desc(padj_pairwise), desc(p.value_pairwise)),
  aes(
    x = pmax(expression.trend_shrunk_1, -.035),
    y = expression.trend_shrunk_2
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey50"
  ) +
  rasterize(geom_point(
    aes(
      color = direction_pairwise_sign,
      text = gene,
      shape = expression.trend_shrunk_1 < -.035
    ),
    alpha = .6,
    size = 2
  ), dpi = 300, dev = "ragg_png") +
  geom_text_repel(
    data = \(df) {
      # browser()
      arrange(df, padj_pairwise, p.value_pairwise) |>
      group_by(pert_name_1, pert_name_2, direction_1, direction_2) |>
      mutate(
        gene_label = if_else(
          seq_len(n()) <= 10 & padj_pairwise < 0.05,
          gene, ""
        )
      ) |>
      ungroup()
    },
    # data = \(df) {
    #   arrange(df, padj_pairwise, p.value_pairwise) |>
    #     group_by(pert_name_1, pert_name_2, direction_combined) |>
    #     filter(padj_pairwise < 0.05) |>
    #     slice_head(n = 10) |>
    #     ungroup() |>
    #     mutate(
    #       gene_label = gene
    #     )
    # },
    aes(label = gene_label),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e7,
    seed = 42,
    # min.segment.length = 0,
  ) +
  scale_shape_manual(
    values = c(
      `TRUE` = "◀",
      `FALSE` = "●"
    ),
    guide = "none"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(.1, .1))
  ) +
  scale_color_manual(
    values = c(
      "positive" = paletteer::paletteer_c("ggthemes::Sunset-Sunrise Diverging", 10)[10],
      "negative" =  paletteer::paletteer_c("ggthemes::Sunset-Sunrise Diverging", 10)[1],
      "not significant" = "black"
    ),
    guide = guide_legend(
      override.aes = list(
        shape = 16
      )
    )
  ) +
  coord_equal() +
  labs(
    x = "AOC vs gene expression\nYNW-10-135",
    y = "AOC vs gene expression\nYKL-5-124"
  )

library(Cairo)

CairoPDF(
  file.path(plot_dir, "slopes_comparison_cairo.pdf"),
  width = 4, height = 4
)
print(p)
dev.off()
ggsave(
  file.path(plot_dir, "slopes_comparison.pdf"),
  p, width = 4, height = 4,
  device = cairo_pdf
)

plotly::ggplotly(p, tooltip = "text")
```

## Compare with and without MDR filtering

```{r}
model_slope_comp <- power_inner_join(
  model_slopes |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(padj)
    ),
  model_mdr_included_slopes |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(padj)
    ),
  by = c("gene", "pert_name"),
  suffix = c("_filtered", "_unfiltered"),
  check = check_specs(
    unmatched_keys_left = "warn",
    unmatched_keys_right = "warn"
  )
)

p <- ggplot(
  model_slope_comp,
  aes(
    x = signed_p_unfiltered,
    y = signed_p_filtered
  )
) +
  geom_point(
    alpha = .6,
    shape = 16,
    color = "#aaaaaa"
  ) +
  geom_abline(
    linetype = "dashed",
    slope = 1, intercept = 0
  ) +
  geom_text_repel(
    data = \(x) bind_rows(
      x |> arrange(padj_filtered) |>
        filter(padj_filtered < 0.05) |>
        mutate(
          direction = if_else(
            direction_filtered == "negative", "negative", "positive"
          )
        ) |>
        group_by(pert_name, direction) |>
        slice_head(n = 5),
      x |> arrange(padj_unfiltered) |>
        filter(padj_unfiltered < 0.05) |>
        mutate(
          direction = if_else(
            direction_unfiltered == "negative", "negative", "positive"
          )
        ) |>
        group_by(pert_name, direction) |>
        slice_head(n = 5)
    ) |> distinct(),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e6,
    seed = 42,
    min.segment.length = 0
  ) +
  facet_wrap(
    ~pert_name,
    axes = "all", axis.labels = "margins",
    scales = "free"
  ) +
  labs(
    x = "Signed -log10(adj. p-value) without MDR filtering",
    y = "Signed -log10(adj. p-value) with MDR filtering"
  )

ggsave(
  file.path(plot_dir, "slope_p_value_comparison.pdf"),
  p, width = 10, height = 4
)
```


## Enrichment analysis


```{r}
database_abbreviations <- c(
  "REACTOME" = "R",
  "HALLMARK" = "H",
  "PID" = "P",
  "KEGG_MEDICUS" = "K",
  "GOBP" = "BP",
  "GOMF" = "MF",
  "GOCC" = "CC"
)

abbreviate_prefix <- function(strings, abbr_map = database_abbreviations) {
  # Create regex pattern from abbreviation names
  pattern <- paste0("^(", paste(names(abbr_map), collapse = "|"), ")_")

  # Replace matching prefixes with abbreviations
  str_replace(strings, pattern, function(x) {
    prefix <- str_remove(x, "_$")
    paste0(abbr_map[prefix], "_")
  })
}

all_gene_sets <- msigdbr(species = "Homo sapiens")
selected_gene_sets <- all_gene_sets %>%
  filter(
    gs_collection %in% c("H") |
      gs_subcollection %in% c(
        "CP:REACTOME",
        "CP:KEGG_MEDICUS"
      ),
    !str_detect(gs_name, coll("MEDICUS_PATHOGEN")),
    !str_detect(gs_name, coll("MEDICUS_VARIANT"))
  ) %>%
  mutate(
    database = str_extract(gs_name, paste0("^(", paste(names(database_abbreviations), collapse = "|"), ")")),
    pathway = abbreviate_prefix(gs_name)
  )

selected_gene_sets_list <- selected_gene_sets %>%
  group_nest(gs_collection, gs_subcollection, gs_name, database, pathway) %>%
  mutate(
    gs = map(
      data,
      \(x) unique(x$gene_symbol)
    ) %>%
      set_names(pathway)
  )

fora_input_raw <- tribble(
  ~model_type, ~de_type, ~data,
  "linear", "slopes", model_slopes |>
    filter(padj < 0.05) |>
    arrange(p.value) |>
    group_by(term = pert_name, direction) |>
    slice_head(n = 500),
  "linear", "pairwise", model_res_pairwise |>
    filter(padj < 0.05) |>
    arrange(p.value) |>
    group_by(term = contrast, direction) |>
    slice_head(n = 500),
  # "rank", "slopes", model_slopes_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = pert_name, direction) |>
  #   slice_head(n = 500),
  # "rank", "pairwise", model_res_pairwise_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = contrast, direction) |>
  #   slice_head(n = 500)
) |>
  mutate(
    across(data, \(x) map(x, ungroup))
  ) |>
  unnest(data) |>
  nest(.by = c(model_type, de_type, term, direction))

gene_universe <- names(eligible_genes)[eligible_genes]

fora_res_raw <- fora_input_raw %>%
  mutate(
    res = map(
      data,
      \(x) {
        withr::with_seed(42, fora(
          selected_gene_sets_list$gs,
          x$gene,
          universe = gene_universe
        ))
      }
    ),
    res_reduced = future_pmap(
      list(res, data),
      \(res, data) {
        collapsePathwaysORA(
          filter(res, padj < .05),
          selected_gene_sets_list$gs,
          data$gene,
          gene_universe
        )
      },
      .progress = TRUE
    )
  )

qs_save(
  fora_res_raw,
  file.path(result_dir, "fora_res_raw.qs")
)

fora_res_long <- fora_res_raw %>%
  transmute(
    model_type, de_type, term, direction,
    res = map2(
      res, res_reduced,
      \(full, reduced) {
        mutate(
          full,
          is_main = pathway %in% reduced$mainPathways,
          parent_pathway = reduced$parentPathways[pathway]
        )
      }
    )
  ) %>%
  unnest(res) |>
  mutate(
    signed_padj = if_else(
      # multiplied by -1
      direction == "negative", 1, -1
    ) * log10(padj)
  )

write_csv(
  fora_res_long,
  file.path(result_dir, "fora_res_long.csv.gz")
)
```

```{r}
fora_res_top <- fora_res_long |>
  filter(padj < 0.05) |>
  arrange(padj, pval) |>
  group_by(model_type, de_type, term, direction) |>
  slice_head(n = 20) |>
  ungroup() |>
  arrange(padj) |>
  mutate(
    padj_cut = cut(
      padj,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("<0.001", "<0.01", "<0.05", "ns")
    ) |>
      fct_rev(),
    logEnrichment = log10(foldEnrichment) * if_else(direction == "negative", -1, 1),
    pathway_display = str_replace_all(pathway, coll("_"), " ")
  )


ps <- group_nest(
  fora_res_top, model_type, term
) |>
  mutate(
    p = map(
      data,
      \(df) {
        abs_max_est_perc <- max(abs(quantile(df$signed_padj, c(0.025, .975))), na.rm = TRUE)
        ggplot(
          df |>
            mutate(across(pathway_display, \(x) fct_reorder(x, logEnrichment))),
          aes(
            x = logEnrichment,
            y = pathway_display
          )
        ) +
          geom_vline(xintercept = 0, linetype = "dashed") +
          geom_point(
            aes(
              fill = signed_padj,
              size = padj_cut
            ),
            shape = 21,
            color = "black"
          ) +
          paletteer::scale_fill_paletteer_c(
            "ggthemes::Orange-Blue-White Diverging",
            limits = c(-abs_max_est_perc, abs_max_est_perc),
            oob = scales::squish
          ) +
          labs(
            x = "log10(Fold Enrichment)", y = NULL,
            fill = "log10(adj. p)",
            size = "Adjusted p"
          ) +
          envalysis::theme_publish()
          # scale_x_log10()
      }
    )
  )

dir.create(
  file.path(plot_dir, "enrichment"),
  showWarnings = FALSE, recursive = TRUE
)
pwalk(
  ps,
  \(term, p, model_type, ...) {
    ggsave(
      file.path(plot_dir, "enrichment", paste0(model_type, "_", term, "_enrichment.pdf")),
      p, width = 12, height = 7
    )
  }
)
```

### FGSEA

```{r}

fgsea_input_raw <- tribble(
  ~model_type, ~de_type, ~data,
  "linear", "slopes", model_slopes |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(p.value)
    ) |>
    group_by(term = pert_name) |>
    summarize(
      ranks = list(set_names(signed_p, gene)),
      .groups = "drop"
    ),
  "linear", "pairwise", model_res_pairwise |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(p.value)
    ) |>
    group_by(term = contrast) |>
    summarize(
      ranks = list(set_names(signed_p, gene)),
      .groups = "drop"
    )
  # "rank", "slopes", model_slopes_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = pert_name, direction) |>
  #   slice_head(n = 500),
  # "rank", "pairwise", model_res_pairwise_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = contrast, direction) |>
  #   slice_head(n = 500)
) |>
  mutate(
    across(data, \(x) map(x, ungroup))
  ) |>
  unnest(data)

plan(multicore, workers = 6)
fgsea_res_raw <- fgsea_input_raw %>%
  mutate(
    res_raw = future_map(
      ranks,
      \(x) {
        withr::with_seed(42, purrr::quietly(fgseaMultilevel)(
          selected_gene_sets_list$gs,
          x,
          nPermSimple = 100000
        ))
      },
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    ),
    res = map(res_raw, "result"),
    res_reduced = future_pmap(
      list(res, ranks),
      \(res, x) {
        collapsePathways(
          filter(res, padj < .05),
          selected_gene_sets_list$gs,
          x
        )
      },
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    ),
    warnings = map_chr(res_raw, \(x) paste(x$warnings, collapse = "; "))
  )

qs_save(
  fgsea_res_raw,
  file.path(result_dir, "fgsea_res_raw.qs")
)

fgsea_res_long <- fgsea_res_raw %>%
  transmute(
    model_type, de_type, term,
    res = map2(
      res, res_reduced,
      \(full, reduced) {
        mutate(
          full,
          is_main = pathway %in% reduced$mainPathways,
          parent_pathway = reduced$parentPathways[pathway]
        )
      }
    )
  ) %>%
  unnest(res) |>
  mutate(
    direction = if_else(ES < 0, "negative", "positive"),
    signed_p = if_else(
      direction == "negative", 1, -1
    ) * log10(pval)
  )

write_csv(
  fgsea_res_long,
  file.path(result_dir, "fgsea_res_long.csv.gz")
)
```


```{r}
fgsea_res_top <- fgsea_res_long |>
  filter(padj < 0.05) |>
  arrange(padj, pval) |>
  group_by(model_type, de_type, term, direction) |>
  slice_head(n = 20) |>
  ungroup() |>
  arrange(padj) |>
  mutate(
    padj_cut = cut(
      padj,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("<0.001", "<0.01", "<0.05", "ns")
    ) |>
      fct_rev(),
    pathway_display = str_replace_all(pathway, coll("_"), " ")
  )


ps <- group_nest(
  fgsea_res_top, model_type, term
) |>
  mutate(
    p = map(
      data,
      \(df) {
        abs_max_est_perc <- max(abs(quantile(df$signed_p, c(0.025, .975))), na.rm = TRUE)
        ggplot(
          df |>
            mutate(across(pathway_display, \(x) fct_reorder(str_trunc(x, 50,), NES))),
          aes(
            x = NES,
            y = pathway_display
          )
        ) +
          geom_vline(xintercept = 0, linetype = "dashed") +
          geom_point(
            aes(
              fill = signed_p,
              size = padj_cut,
              alpha = is_main
            ),
            shape = 21,
            color = "black"
          ) +
          scale_alpha_manual(
            values = c(
              `TRUE` = 1,
              `FALSE` = 0.2
            ),
            guide = "none"
          ) +
          paletteer::scale_fill_paletteer_c(
            "ggthemes::Orange-Blue-White Diverging",
            limits = c(-abs_max_est_perc, abs_max_est_perc),
            oob = scales::squish
          ) +
          labs(
            x = "Normalized enrichment score", y = NULL,
            fill = "log10(adj. p)",
            size = "Adjusted p"
          )
          # scale_x_log10()
      }
    )
  )

dir.create(
  file.path(plot_dir, "enrichment"),
  showWarnings = FALSE, recursive = TRUE
)
pwalk(
  ps,
  \(term, p, model_type, ...) {
    ggsave(
      file.path(plot_dir, "enrichment", paste0("fgsea_", model_type, "_", term, "_enrichment.pdf")),
      p, width = 8, height = 8
    )
  }
)
```

#### Dot plot comparison

```{r}
library(seriation)

cluster_df <- function(df, row_var, col_var, value_var, values_fill = 0) {
  # browser()
  mat <- df %>%
    select({{row_var}}, {{col_var}}, {{value_var}}) %>%
    pivot_wider(names_from = {{col_var}}, values_from = {{value_var}}, values_fill = values_fill) %>%
    column_to_rownames(rlang::as_name(rlang::enquo(row_var)))
  # browser()
  if (rlang::is_bare_numeric(pull(df, {{value_var}}))) {
    dist_rows <- dist(mat, method = "euclidian")
    dist_cols <- dist(t(mat), method = "euclidian")
  } else {
    # browser()
    dist_rows <- cluster::daisy(mat, metric = "gower")
    dist_cols <- t(mat) %>%
      as.data.frame() %>%
      mutate(across(everything(), \(x) factor(x, levels = levels(pull(df, {{value_var}}))))) %>%
      cluster::daisy(metric = "gower")
  }
  clust_rows <- hclust(dist_rows, method = "average") %>%
      reorder(dist_rows, method = "olo")
  clust_cols <- hclust(dist_cols, method = "average") %>%
      reorder(dist_cols, method = "olo")
  df %>%
    mutate(
      "{{row_var}}" := factor({{row_var}}, levels = clust_rows$labels[clust_rows$order]),
      "{{col_var}}" := factor({{col_var}}, levels = clust_cols$labels[clust_cols$order])
    )
}
```

```{r}
fgsea_res_long_relevant <- fgsea_res_long |>
  filter(
    model_type == "linear",
    term %in% c("YKL-5-124", "YNW-10-135"),
    !str_detect(pathway, coll("SARS")),
    !str_detect(pathway, coll("INFLUENZA")),
  )

fgsea_res_top_comp <- fgsea_res_long_relevant |>
  semi_join(
    fgsea_res_long_relevant |>
      group_by(term) |>
      arrange(padj, pval) |>
      filter(
        padj < 0.05,
        is_main
      ) |>
      slice_head(n = 20),
    by = "pathway"
  ) |>
  mutate(
    padj_cut = cut(
      padj,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("<0.001", "<0.01", "<0.05", "ns")
    ) |>
      fct_rev(),
    pathway_display = str_replace_all(pathway, coll("_"), " ") |>
      str_sub(start = 3) |>
      str_wrap(width = 50)
      # str_trunc(width = 60)
  ) |>
  drop_na(signed_p) |>
  cluster_df(
    pathway_display, term, signed_p
  )


p <- ggplot(
  fgsea_res_top_comp,
  aes(
    x = term,
    y = pathway_display
  )
) +
  geom_point(
    aes(
      color = NES,
      size = padj_cut
    ),
    shape = 16,
    alpha = 1
  ) +
  paletteer::scale_color_paletteer_c(
    "ggthemes::Red-Blue Diverging",
    limits = c(-max(abs(fgsea_res_top_comp$NES)), max(abs(fgsea_res_top_comp$NES))),
    oob = scales::squish,
    direction = -1
  ) +
  scale_size_manual(
    values = c(
      "<0.001" = 3,
      "<0.01" = 1.5,
      "<0.05" = 1,
      "ns" = 0.5
    )
  ) +
  labs(
    x = NULL, y = NULL,
    color = "Normalized enrichment score",
    size = "Adjusted p"
  ) +
  scale_x_discrete(position = "top") +
  theme_minimal(base_size = 10) +
  theme(
    text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 0, vjust = 0),
    panel.grid.major.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(
      color = "black",
      lineheight = .65
    ),
    legend.position = "bottom",
    legend.box = "vertical"
  )

p

ggsave(
  file.path(plot_dir, "enrichment", "fgsea_slopes_comparison_ykl_5_124_vs_ynw_10_135.pdf"),
  p, width = 4.2, height = 7.5
)

# plotEnrichment(
#   selected_gene_sets_list$gs[["R_RRNA_PROCESSING"]],
#   fgsea_input_raw$ranks[[3]]
# )

# fgsea_res_long_relevant |>
#   filter(pathway == "R_RRNA_PROCESSING")

# fgseaMultilevel(
#   selected_gene_sets_list$gs["R_RRNA_PROCESSING"],
#   fgsea_input_raw$ranks[[3]],
#   nPermSimple = 1000000
# )
```


```{r}

fgsea_res_long_relevant <- fgsea_res_long |>
  filter(
    model_type == "linear",
    term %in% c("YKL-5-124", "YNW-10-135", "(YKL-5-124) - (YNW-10-135)"),
    !str_detect(pathway, coll("SARS")),
    !str_detect(pathway, coll("INFLUENZA")),
  )

fgsea_res_top_comp <- fgsea_res_long_relevant |>
  semi_join(
    fgsea_res_long_relevant |>
      group_by(term) |>
      arrange(padj, pval) |>
      filter(
        padj < 0.05,
        is_main
      ) |>
      slice_head(n = 20),
    by = "pathway"
  ) |>
  mutate(
    padj_cut = cut(
      padj,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("<0.001", "<0.01", "<0.05", "ns")
    ) |>
      fct_rev(),
    pathway_display = str_replace_all(pathway, coll("_"), " ") |>
      str_sub(start = 3) |>
      str_wrap(width = 50)
      # str_trunc(width = 60)
  ) |>
  drop_na(signed_p) |>
  cluster_df(
    pathway_display, term, signed_p
  )


p <- ggplot(
  fgsea_res_top_comp,
  aes(
    x = term,
    y = pathway_display
  )
) +
  geom_point(
    aes(
      color = NES,
      size = padj_cut
    ),
    shape = 16,
    alpha = 1
  ) +
  paletteer::scale_color_paletteer_c(
    "ggthemes::Red-Blue Diverging",
    limits = c(-max(abs(fgsea_res_top_comp$NES)), max(abs(fgsea_res_top_comp$NES))),
    oob = scales::squish,
    direction = -1
  ) +
  scale_size_manual(
    values = c(
      "<0.001" = 3,
      "<0.01" = 1.5,
      "<0.05" = 1,
      "ns" = 0.5
    )
  ) +
  labs(
    x = NULL, y = NULL,
    color = "Normalized enrichment score",
    size = "Adjusted p"
  ) +
  scale_x_discrete(position = "top") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 0, vjust = 0),
    panel.grid.major.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(
      lineheight = .65
    ),
    legend.position = "bottom",
    legend.box = "vertical"
  )

p

ggsave(
  file.path(plot_dir, "enrichment", "fgsea_slopes_comparison_ykl_5_124_vs_ynw_10_135.pdf"),
  p, width = 4.2, height = 7.5
)
```


## Make Excel sheet

```{r}
format_p <- function(wb, sheet_name, col_name, table) {
  data_rows <- nrow(table) + 1  # +1 for header
  col_idx <- which(names(table) == col_name)

  conditionalFormatting(
    wb = wb,
    sheet = sheet_name,
    cols = col_idx,
    rows = 2:data_rows,  # Skip header row
    type = "expression",
    rule = paste0(
      "AND(NOT(ISBLANK(", openxlsx::int2col(col_idx), "2)),",
      "ISNUMBER(", openxlsx::int2col(col_idx), "2),",
      openxlsx::int2col(col_idx), "2<0.05)"
    ),
    style = createStyle(bgFill = "#FF9999")  # Light red background
  )
}

format_phenotype <- function(wb, sheet_name, col_name, table, percentile = .001) {
  data_rows <- nrow(table) + 1  # +1 for header
  col_idx <- which(names(table) == col_name)

  min_max <- quantile(table[[col_name]], probs = c(percentile, 1 - percentile), na.rm = TRUE) |>
    abs() |>
    max()

  conditionalFormatting(
    wb = wb,
    sheet = sheet_name,
    cols = col_idx,
    rows = 2:data_rows,  # Skip header row
    type = "colourScale",
    style = c("blue", "white", "red"),
    rule = c(-1, 0, 1) * min_max
  )
}

add_sheet <- function(wb, sheet_name, table) {
  addWorksheet(wb, sheet_name)
  # table <- select(table, where(negate(is.list)))
  writeDataTable(wb, sheet_name, table, tableStyle = "TableStyleLight1")
  walk(
    str_subset(names(table), "^padj"),
    \(col_name) format_p(wb, sheet_name, col_name, table)
  )
  walk(
    str_subset(names(table), "^estimate|^expression.trend|^foldEnrichment|^signed_padj"),
    \(col_name) format_phenotype(wb, sheet_name, col_name, table)
  )
  setColWidths(wb, sheet_name, cols = seq_len(ncol(table)), widths = "auto")
}

wb <- createWorkbook()

list(
  "Slopes" = model_slopes |>
    arrange(padj, p.value),
  "Pairwise" = model_res_pairwise |>
    select(-term) |>
    arrange(padj, p.value)
) |>
  iwalk(\(t, n) add_sheet(wb, n, t))

saveWorkbook(
  wb,
  file.path(result_dir, "cdk7_prism_expression_model.xlsx"),
  overwrite = TRUE
)

wb <- createWorkbook()

fora_res_long |>
  arrange(padj, pval) |>
  relocate(signed_padj, .after = padj) |>
  relocate(overlapGenes, .after = last_col()) |>
  group_by(term) |>
  group_walk(\(df, key) {
    add_sheet(wb, key$term, df)
  })

saveWorkbook(
  wb,
  file.path(result_dir, "cdk7_prism_expression_enrichment.xlsx"),
  overwrite = TRUE
)
```

* Color points in Volcano by direction
* Plot comparison between slopes separately for both compounds, slope in compound
  vs slope difference YKL vs YNW
