---
title: "CDK7 PRISM analysis"
author: "Clemens Hug"
date: "2026-02-06"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(powerjoin)
library(synExtra)
library(conflicted)
library(hdf5r)
library(ggbeeswarm)
library(here)
library(lmerTest)
library(emmeans)
library(broom.mixed)
library(furrr)
library(qs2)
library(ggrepel)
library(msigdbr)
library(fgsea)
library(openxlsx)

conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  base::setdiff,
  lmerTest::lmer
)

synapser::synLogin()
syn <- synDownloader("~/data", .cache = TRUE)

plot_dir <- here("plots", "prism")
dir.create(plot_dir, showWarnings = FALSE)

result_dir <- here("prism")
dir.create(result_dir, showWarnings = FALSE)

set_theme(
  envalysis::theme_publish(
    base_family = "Helvetica",
    base_size = 10
  )
)
```

PRISM analysis code at https://github.com/theprismlab/sushi/blob/5320078b788fd606ec562ac60ebdb4c9688087f5/scripts/biomarker/biomarker_functions.R#L103

## Load Biomarker data

```{r}
biomarkers_raw <- syn("syn73615969") |>
  H5File$new(mode = "r")

expression_col_meta <- biomarkers_raw[["Expression/column_meta"]]$read() |>
  as_tibble()

expression_row_meta <- biomarkers_raw[["Expression/row_meta"]]$read() |>
  as_tibble()

expression_mat <- biomarkers_raw[["Expression/mat"]]$read() |>
  magrittr::set_colnames(
    expression_col_meta$column_name
  ) |>
  magrittr::set_rownames(
    expression_row_meta$ModelID
  )

cell_line_meta <- biomarkers_raw[["Lineage_Table/table"]]$read() |>
  as_tibble()

# for mapping old YKL ccle names to new depmap ids
cell_line_meta_depmap <- syn("syn73674394") |>
  read_csv()
```

## To Do

```{r}
prism_raw_zips <- c(
  mts_30 = "syn73616950",
  ykl = "syn73616961"
) |>
  map(syn)

prism_drc <- list(
  mts_30 = prism_raw_zips[["mts_30"]] |>
    unz("drc/MTS030_TINGHU_ZHANG_DRC_TABLE.csv") |>
    read_csv(),
  ykl = prism_raw_zips[["ykl"]] |>
    unz("data/DRC_TABLE.csv") |>
    read_csv() |>
    rename(
      minimum_dose = min_dose,
      maximum_dose = max_dose,
      inflection = ec50,
      log2_ic50 = log2.ic50,
      frac_var_explained = R2
    ) |>
    power_inner_join(
      distinct(cell_line_meta_depmap, ccle_name = CCLEName, depmap_id = ModelID) |>
        drop_na(),
      by = "ccle_name",
      check = check_specs(
        unmatched_keys_left = "warn",
        duplicate_keys_right = "warn"
      )
    )
) |>
  bind_rows(
    .id = "dataset"
  )


```


## Explore MDR genes

```{r}
mdr_genes <- c("ABCB1", "ABCG2", "ABCC1")

mdr_expression <- expression_mat[, mdr_genes] |>
  as_tibble(rownames = "depmap_id") |>
  pivot_longer(
    -depmap_id,
    names_to = "gene",
    values_to = "expression"
  ) |>
  power_left_join(
    cell_line_meta,
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  )

p <- ggplot(
  mdr_expression |>
    mutate(
      across(lineage, fct_rev)
    ),
  aes(
    x = expression,
    y = lineage,
    color = lineage
  )
) +
  geom_quasirandom(
    shape = 16
  ) +
  facet_wrap(
    ~gene,
    scales = "free_x",
    axes = "all", axis.labels = "margins"
  ) +
  guides(color = "none") +
  scale_x_continuous(
    expand = expansion(mult = c(0.01, .05))
  ) +
  labs(
    x = "Expression", y = NULL
  )
p

ggsave(
  file.path(plot_dir, "mdr_genes_expression.pdf"),
  p, width = 10, height = 6.5
)
```

Check correlation of the three

A sensible cutoff for ABCB1 seems to be around 1-2. Clear negative population with
long positive tail. For ABCG2 bit harder to define. Much fatter tail especially
for lung and skin. ABCC1 doesn't have negative population at all.

Just doing ABCB1 cutoff 1 for now.

## Filter cell lines and genes

```{r}
ABCB1_CUTOFF <- 1

cell_lines_n_profiled <- prism_drc |>
  distinct(depmap_id, pert_name) |>
  count(depmap_id)

cell_lines_n_profiled |>
  count(n)

cell_lines_filtered <- cell_line_meta |>
  semi_join(
    mdr_expression |>
      filter(gene == "ABCB1", expression <= ABCB1_CUTOFF),
    by = "depmap_id"
  ) |>
  semi_join(
    cell_lines_n_profiled |>
      filter(n >= 2),
    by = "depmap_id"
  )


```

Selecting genes with sufficient expression in at least some cell lines.

```{r}
eligible_genes <- apply(
  expression_mat, 2,
  \(x) quantile(x, .8, names = FALSE)
) > 1

```

## Compound correlation

How well are AUC values correlated between the three compounds?

```{r}
auc_comparison <- combn(
  unique(prism_drc$pert_name), 2
) |>
  t() |>
  magrittr::set_colnames(c("compound_1", "compound_2")) |>
  as_tibble() |>
  inner_join(
    select(prism_drc, depmap_id, pert_name, auc),
    by = c("compound_1" = "pert_name"),
    relationship = "many-to-many"
  ) |>
  inner_join(
    select(prism_drc, depmap_id, pert_name, auc),
    by = c("compound_2" = "pert_name", "depmap_id"),
    suffix = c("_1", "_2"),
    relationship = "many-to-many"
  ) |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  )

p <- ggplot(
  auc_comparison,
  aes(
    x = 1 - auc_1,
    y = 1 - auc_2,
    color = lineage
  )
) +
  geom_abline(
    slope = 1, intercept = 0,
    linetype = "dashed",
    color = "grey"
  ) +
  geom_point(
    shape = 16,
    alpha = .6
  ) +
  geom_text(
    aes(
      label = r
    ),
    data = \(x) group_by(x, compound_1, compound_2) |>
      drop_na(auc_1, auc_2) |>
      summarize(
        r = paste0("  R = ", signif(cor(auc_1, auc_2), 2)),
        .groups = "drop"
      ),
    color = "black",
    x = -Inf, y = Inf,
    hjust = 0, vjust = 1
  ) +
  facet_wrap(
    ~paste(compound_1, "vs", compound_2),
    axes = "all", axis.labels = "margins"
  ) +
  labs(
    x = "AOC Compound 1", y = "AOC Compound 2"
  ) +
  coord_equal()
p

ggsave(
  file.path(plot_dir, "auc_comparison.pdf"),
  p, width = 10, height = 7
)
```

## Fitting lineage model

Removing lineages that don't have data for all three compounds.
Otherwise the model won't be able to fit the interaction term properly.
Only affects few cells lines, Adrenal gland, Eye, and Vulva/Vagina.

```{r}
prism_drc_lineage <- prism_drc |>
  select(depmap_id, pert_name, auc) |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  group_by(lineage) |>
  filter(
    n_distinct(pert_name) == 3
  ) |>
  ungroup()

lineage_vector <- cell_line_meta |>
  semi_join(
    prism_drc_lineage,
    by = "depmap_id"
  ) |>
  group_nest(lineage) |>
  mutate(
    gs = set_names(
      map(data, "depmap_id"),
      lineage
    )
  )

prism_fgsea_res <- prism_drc_lineage |>
  drop_na(auc) |>
  group_by(pert_name, depmap_id) |>
  summarize(
    auc = mean(auc[is.finite(auc)], na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(pert_name) |>
  summarize(
    res = fgseaMultilevel(
      lineage_vector$gs,
      set_names(scale(auc)[, 1], depmap_id)
    ) |>
      arrange(padj, pval) |>
      list()
  )
```

```{r}
prism_lineage_model <- lm(
  auc ~ lineage * pert_name,
  data = prism_drc_lineage |>
    filter(depmap_id %in% cell_lines_filtered$depmap_id)
)
car::Anova(prism_lineage_model)

lineage_emm <- emmeans(
  prism_lineage_model,
  ~ lineage | pert_name
)

lineage_emm_res <- contrast(lineage_emm, method = "eff", adjust = "fdr") |>
  as_tibble() |>
  arrange(pert_name, p.value)
```


```{r}
prism_lineage_model_sans_filtering <- lm(
  auc ~ lineage * pert_name,
  data = prism_drc_lineage
)
car::Anova(prism_lineage_model_sans_filtering)

lineage_emm_sans_filtering <- emmeans(
  prism_lineage_model_sans_filtering,
  ~ lineage | pert_name
)

lineage_emm_res_sans_filtering <- contrast(lineage_emm_sans_filtering, method = "eff", adjust = "fdr") |>
  as_tibble() |>
  arrange(pert_name, p.value)
```


```{r}
p <- ggplot(
  prism_drc_lineage |>
    mutate(
      across(lineage, fct_rev),
      filtered = if_else(
        depmap_id %in% cell_lines_filtered$depmap_id,
        "Included",
        "Excluded"
      )
    ),
  aes(
    x = 1 - auc,
    y = lineage,
    color = lineage
  )
) +
  geom_quasirandom(
    aes(
      color = filtered,
      fill = lineage
    ),
    shape = 21
  ) +
  scale_color_manual(
    values = c(
      "Included" = alpha("black", .0),
      "Excluded" = alpha("red", .8)
    ),
    name = "MDR filtering"
  ) +
  geom_text(
    aes(
      label = paste0(
        if_else(estimate > 0, "  ", ""),
        cut(p.value, c(-Inf, 0.001, 0.01, 0.05, Inf), c("***", "**", "*", "")),
        if_else(estimate > 0, "", "  ")
      ),
      x = if_else(estimate > 0, -Inf, Inf),
      hjust = if_else(estimate > 0, 0, 1),
    ),
    data = lineage_emm_res_sans_filtering |>
      mutate(
        lineage = str_remove(contrast, " effect") |>
          str_remove_all("\\(|\\)")
      ),
    color = "black",
    vjust = .5
  ) +
  scale_fill_discrete(
    palette = \(n) scales::alpha(scales::hue_pal()(n), .6),
    guide = "none"
  ) +
  ggnewscale::new_scale_color() +
  geom_boxplot(
    aes(
      color = lineage
    ),
    outlier.shape = NA,
    fill = NA
  ) +
  scale_color_discrete(
    guide = "none"
  ) +
  facet_wrap(
    ~pert_name,
    scales = "free_x",
    axes = "all", axis.labels = "margins"
  ) +
  labs(
    x = "Area over curve (AOC)", y = NULL,
    color = ""
  )
p

ggsave(
  file.path(plot_dir, "auc_by_lineage.pdf"),
  p, width = 8, height = 4.5
)
```

## Fitting expression model

The goal is to fit a linear model of AUC ~ expression * drug

```{r}
expression_mat_ifit_composite <- cbind(
  expression_mat,
  expression_mat[, c("IFITM1", "IFITM2", "IFITM3")] |>
    scale() |>
    apply(1, mean) |>
    as.matrix(ncol = 1) |>
    magrittr::set_colnames("IFITM-composite")
)

input_data <- prism_drc |>
  filter(
    depmap_id %in% cell_lines_filtered$depmap_id
  ) |>
  select(depmap_id, pert_name, auc)  |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  power_left_join(
    expression_mat_ifit_composite[
      cell_lines_filtered$depmap_id,
      c(eligible_genes, c("IFITM-composite" = TRUE))
    ] |>
      as_tibble(rownames = "depmap_id") |>
      pivot_longer(
        -depmap_id,
        names_to = "gene",
        values_to = "expression"
      ),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn"
    )
  ) |>
  group_nest(gene)
```

```{r}
run_model <- function(
  df, model_formula, random_effects = TRUE,
  lmer.df = "satterthwaite", REML = FALSE,
  rank_transform = FALSE
) {
  # browser()
  if (rank_transform) {
    df <- df |>
      mutate(
        across(
          c(auc, expression),
          rank
        )
      )
  }
  if (random_effects) {
    model <- lmer(
      formula(model_formula),
      data = df,
      REML = REML
    )
  } else {
    model <- lm(
      formula(model_formula),
      data = df
    )
  }

  emm <- emtrends(
    model,
    ~pert_name,
    var = "expression",
    adjust = "none",
    lmer.df = lmer.df
  )

  emm_intercepts <- emmeans(
    model,
    ~pert_name,
    at = list(expression = 0),
    adjust = "none",
    lmer.df = lmer.df
  )

  res_by_compound <- inner_join(
    tidy(emm_intercepts) |>
      rename(intercept = estimate) |>
      select(-c(df, statistic, p.value)),
    tidy(emm),
    by = "pert_name",
    suffix = c("_intercept", "_slope")
  )

  list(
    model_stats = glance(model),
    model_coefs = tidy(model),
    model_by_compound = res_by_compound,
    pairwise_res = pairs(
      emm,
      adjust = "none"
    ) |>
      tidy()
  )
}

```

```{r}
model_formulas <- tribble(
  ~model_name, ~model_formula, ~random_effects,
  "random_lineage", auc ~ expression * pert_name + (1 | lineage), TRUE,
  "random_cell_line", auc ~ expression * pert_name + (1 | depmap_id), TRUE,
  "fixed", auc ~ expression * pert_name, FALSE
)

test_output <- input_data |>
  slice(1) |>
  crossing(model_formulas) |>
  mutate(
    model_out = pmap(
      list(data, model_formula, random_effects),
      \(d, f, r) run_model(
        d,
        f,
        r,
        REML = FALSE
      )
    )
  )

plan(multicore, workers = 4)

model_res_raw_all <- input_data |>
  crossing(model_formulas) |>
  mutate(
    model_out = future_pmap(
      list(data, model_formula, random_effects),
      \(data, model_formula, random_effects) run_model(
        data,
        model_formula,
        random_effects,
        REML = FALSE
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )

qs_save(
  model_res_raw_all,
  file.path(result_dir, "model_res_all_no_reml.qs")
)
# model_res_raw <- qd_read(
#   file.path(result_dir, "model_res_raw.qs")
# )


model_res_aic <- model_res_raw_all |>
  transmute(
    gene, model_name,
    aic = map_dbl(
      model_out, \(x) x$model_stats$AIC
    ),
    bic = map_dbl(
      model_out, \(x) x$model_stats$BIC
    )
  ) |>
  group_by(gene) |>
  mutate(
    rank_aic = rank(-aic),
    rank_bic = rank(-bic)
  ) |>
  ungroup()

model_res_aic_summary <- model_res_aic |>
  group_by(model_name) |>
  summarize(
    n_best_aic = sum(rank_aic == 1),
    n_best_bic = sum(rank_bic == 1),
    mean_aic = mean(rank_aic),
    mean_bic = mean(rank_bic),
    .groups = "drop"
  )

```

  model_name       n_best_aic n_best_bic mean_aic mean_bic
  <chr>                 <int>      <int>    <dbl>    <dbl>
1 fixed                  3852         44     1.71     2.00
2 random_cell_line          0          0     3        3
3 random_lineage         9424      13232     1.29     1.00

Running the random_lineage model

```{r}
plan(multicore, workers = 8)

model_res_raw <- input_data |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger"
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )

qs_save(
  model_res_raw,
  file.path(result_dir, "model_res_raw.qs")
)

model_res_rank_raw <- input_data |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger",
        rank_transform = TRUE
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )

qs_save(
  model_res_rank_raw,
  file.path(result_dir, "model_res_rank_raw.qs")
)
```


```{r}
model_res_pairwise <- model_res_raw |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(estimate > 0, "positive", "negative")
  )

model_slopes <- model_res_raw |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```

```{r}
model_res_pairwise_rank <- model_res_rank_raw |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(estimate > 0, "positive", "negative")
  )

model_slopes_rank <- model_res_rank_raw |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```

### Without MDR filtering

```{r}
cell_lines_sans_mdr_filtering <- cell_line_meta |>
  filter(
    depmap_id %in% rownames(expression_mat)
  ) |>
  semi_join(
    cell_lines_n_profiled |>
      filter(n >= 2),
    by = "depmap_id"
  )

input_data_sans_mdr_filtering <- prism_drc |>
  filter(
    depmap_id %in% cell_lines_sans_mdr_filtering$depmap_id
  ) |>
  select(depmap_id, pert_name, auc)  |>
  power_inner_join(
    cell_line_meta |>
      select(depmap_id, lineage),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) |>
  power_left_join(
    expression_mat_ifit_composite[
      cell_lines_sans_mdr_filtering$depmap_id,
      c(eligible_genes, c("IFITM-composite" = TRUE))
    ] |>
      as_tibble(rownames = "depmap_id") |>
      pivot_longer(
        -depmap_id,
        names_to = "gene",
        values_to = "expression"
      ),
    by = "depmap_id",
    check = check_specs(
      unmatched_keys_left = "warn"
    )
  ) |>
  group_nest(gene)

```


```{r}
plan(multicore, workers = 6)

model_res_raw2 <- input_data_sans_mdr_filtering |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger"
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )
qs_save(
  model_res_raw2,
  file.path(result_dir, "model_res_raw_no_mdr_filtering.qs")
)
# model_res_raw <- qs_read(
#   file.path(result_dir, "model_res_raw_no_mdr_filtering.qs")
# )

model_res_rank_raw <- input_data_sans_mdr_filtering |>
  mutate(
    model_out = future_map(
      data,
      \(data) run_model(
        data,
        auc ~ expression * pert_name + (1 | lineage),
        random_effects = TRUE,
        REML = TRUE,
        lmer.df = "kenward-roger",
        rank_transform = TRUE
      ),
      .progress = TRUE,
      .options = furrr_options(seed = 42)
    )
  )
qs_save(
  model_res_rank_raw,
  file.path(result_dir, "model_res_rank_raw_no_mdr_filtering.qs")
 )
```


* Fraction of cell lines that respond for each compound (AUC < .25)
* Overview of ABCB1 and IFITM expression
* ABCB1/AUC correlation
* Plot correlation IFIT/AUC
* Make composite IFITM score


```{r}
model_mdr_included_res_pairwise_ <- model_res_raw2 |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(estimate > 0, "positive", "negative")
  )

model_mdr_included_slopes <- model_res_raw2 |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```


```{r}
model_res_pairwise_rank <- model_res_rank_raw |>
  transmute(
    gene,
    pairwise_res = map(model_out, "pairwise_res")
  ) |>
  unnest(pairwise_res) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(estimate > 0, "positive", "negative")
  )

model_slopes_rank <- model_res_rank_raw |>
  transmute(
    gene,
    model_slopes = map(model_out, "model_by_compound")
  ) |>
  unnest(model_slopes) |>
  mutate(
    padj = p.adjust(p.value, method = "fdr"),
    direction = if_else(expression.trend > 0, "positive", "negative")
  )
```


## Plot

### P-value distribution

```{r}
p <- bind_rows(
  `Pairwise comparisons` = model_res_pairwise,
  `Model slopes` = model_slopes,
  .id = "result_type"
) |>
  ggplot(
    aes(x = p.value)
  ) +
  geom_histogram() +
  scale_x_continuous(
    expand = expansion(mult = c(0, .05))
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, .05))
  ) +
  facet_wrap(
    ~result_type,
    scales = "free_y",
    axes = "all", axis.labels = "margins"
  ) +
  envalysis::theme_publish()

ggsave(
  file.path(plot_dir, "p_value_distributions.pdf"),
  p, width = 8, height = 4
)
```


```{r}
plot_auc_expression_corr <- function(gene, rank_transform = FALSE, mdr_filtering = TRUE) {
  if (mdr_filtering) {
    id <- input_data
  } else {
    id <- input_data_sans_mdr_filtering
  }
  scatter_df <- id |>
    filter(.data$gene == .env$gene) |>
    unnest(data) |>
    select(-lineage) |>
    power_inner_join(
      cell_line_meta,
      by = "depmap_id",
      check = check_specs(
        unmatched_keys_left = "warn",
        duplicate_keys_right = "warn"
      )
    )
  if (rank_transform) {
    scatter_df <- scatter_df |>
      mutate(
        across(
          c(auc, expression),
          rank
        )
      )
    slope_df <- model_slopes_rank |>
      filter(.data$gene == .env$gene)
  } else {
    if (mdr_filtering) {
      slope_df <- model_slopes |>
        filter(.data$gene == .env$gene)
    } else {
      slope_df <- model_mdr_included_slopes |>
        filter(.data$gene == .env$gene)
    }
  }
  ggplot(
    scatter_df,
    aes(
      x = expression,
      y = 1 - auc
    )
  ) +
    geom_point(
      aes(
        color = lineage
      ),
      alpha = .5,
      shape = 16
    ) +
    geom_abline(
      aes(
        slope = -expression.trend,
        intercept = 1 - intercept
      ),
      data = slope_df
    ) +
    ggnewscale::new_scale_color() +
    geom_text(
      aes(
        label = paste0(
          " b = ", signif(-expression.trend, 2),
          "\n p = ", signif(padj, 2)
        ),
        color = padj < 0.05
      ),
      data = slope_df,
      x = -Inf, y = Inf,
      hjust = 0, vjust = 1
    ) +
    scale_color_manual(
      values = c(
        "TRUE" = "red",
        "FALSE" = "black"
      ),
      guide = "none"
    ) +
    labs(
      x = "Expression", y = "AOC",
      color = "Lineage",
      title = gene
    ) +
    facet_wrap(
      ~pert_name,
      axes = "all", axis.labels = "margins"
    )
}

plot_auc_expression_corr("ABCB1", mdr_filtering = FALSE)
plot_auc_expression_corr("IFITM-composite", mdr_filtering = FALSE)
plot_auc_expression_corr("WDR54")
plot_auc_expression_corr("IFITM1")
plot_auc_expression_corr("AURKB")
plot_auc_expression_corr("CDK7")
plot_auc_expression_corr("MED12L")
plot_auc_expression_corr("PPIA")
plot_auc_expression_corr("IFIT1")
plot_auc_expression_corr("ZNF775")

plot_auc_expression_corr("PAQR3", rank_transform = TRUE)

input_data |>
  filter(gene == "ABCB1") |>
  chuck("data", 1L) |>
  plot_auc_expression_corr()

dir.create(
  file.path(plot_dir, "gene_aoc_correlation"),
  showWarnings = FALSE, recursive = TRUE
)

p <- plot_auc_expression_corr("ABCB1", mdr_filtering = FALSE)
ggsave(
  file.path(plot_dir, "gene_aoc_correlation", "ABCB1.pdf"),
  p, width = 7, height = 4
)

p <- plot_auc_expression_corr("IFITM-composite")
ggsave(
  file.path(plot_dir, "gene_aoc_correlation", "IFITM-composite.pdf"),
  p, width = 7, height = 4
)
```

Make plots for top 20 hits.


```{r}
dir.create(
  file.path(plot_dir, "top_hits", "pairwise"),
  showWarnings = FALSE, recursive = TRUE
)
dir.create(
  file.path(plot_dir, "top_hits", "slopes"),
  showWarnings = FALSE, recursive = TRUE
)

model_res_pairwise |>
  arrange(p.value) |>
  filter(padj < 0.05) |>
  group_by(contrast) |>
  slice_head(n = 20) |>
  mutate(
    rank = seq_len(n())
  ) |>
  ungroup() |>
  pwalk(
    \(gene, contrast, rank, ...) {
      p <- plot_auc_expression_corr(gene)
      ggsave(
        file.path(plot_dir, "top_hits", "pairwise", paste0(contrast, "_", rank, "_", gene, ".pdf")),
        p, width = 8, height = 6
      )
    }
  )

model_slopes |>
  arrange(p.value) |>
  filter(padj < 0.05) |>
  group_by(pert_name) |>
  slice_head(n = 20) |>
  mutate(
    rank = seq_len(n())
  ) |>
  ungroup() |>
  pwalk(
    \(gene, pert_name, rank, ...) {
      p <- plot_auc_expression_corr(gene)
      ggsave(
        file.path(plot_dir, "top_hits", "slopes", paste0(pert_name, "_", rank, "_", gene, ".pdf")),
        p, width = 8, height = 6
      )
    }
  )
```


## Volcano plot

```{r}
p <- ggplot(
  model_slopes |>
    mutate(
      p_censored = -log10(padj) > 12,
      neglog10_padj = pmin(-log10(padj), 12)
    ),
  aes(
    x = expression.trend,
    y = neglog10_padj
  )
) +
  geom_point(
    aes(
      color = padj < 0.05,
      shape = p_censored
    ),
    alpha = .6
  ) +
  geom_text_repel(
    data = \(x) arrange(x, p.value) |>
      filter(padj < 0.05) |>
      mutate(
        direction = if_else(
          expression.trend > 0, "positive", "negative"
        )
      ) |>
      group_by(pert_name, direction) |>
      slice_head(n = 20),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e6,
    seed = 42
  ) +
  scale_shape_manual(
    values = c(
      `TRUE` = 17,
      `FALSE` = 16
    ),
    guide = "none"
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = "red",
      "FALSE" = "black"
    ),
    guide = "none"
  ) +
  facet_wrap(
    ~pert_name,
    axes = "all", axis.labels = "margins",
    scales = "free_y"
  ) +
  labs(
    x = "Slope", y = "-log10(adj. p-value)"
  )
p

ggsave(
  file.path(plot_dir, "volcano_slopes.pdf"),
  p, width = 7, height = 3.5
)
```

```{r}
p <- ggplot(
  model_res_pairwise |>
    mutate(
      p_censored = -log10(padj) > 10,
      neglog10_padj = pmin(-log10(padj), 10)
    ),
  aes(
    x = estimate,
    y = neglog10_padj
  )
) +
  geom_point(
    aes(
      color = padj < 0.05,
      shape = p_censored
    ),
    alpha = .6
  ) +
  geom_text_repel(
    data = \(x) arrange(x, p.value) |>
      mutate(
        direction = if_else(
          estimate > 0, "positive", "negative"
        )
      ) |>
      filter(padj < 0.05) |>
      group_by(contrast, direction) |>
      slice_head(n = 20),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e6,
    seed = 42
  ) +
  scale_shape_manual(
    values = c(
      `TRUE` = 17,
      `FALSE` = 16
    ),
    guide = "none"
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = "red",
      "FALSE" = "black"
    ),
    guide = "none"
  ) +
  facet_wrap(
    ~contrast,
    axes = "all", axis.labels = "margins",
    scales = "free_y"
  ) +
  labs(
    x = "Slope difference", y = "-log10(adj. p-value)"
  )
p

ggsave(
  file.path(plot_dir, "volcano_pairwise.pdf"),
  p, width = 7, height = 3.5
)
```


### Without MDR filtering

```{r}
p <- ggplot(
  model_mdr_included_slopes,
  aes(
    x = expression.trend,
    y = -log10(padj)
  )
) +
  geom_point(
    aes(
      color = padj < 0.05
    ),
    shape = 16,
    alpha = .6
  ) +
  geom_text_repel(
    data = \(x) arrange(x, p.value) |>
      filter(padj < 0.05) |>
      mutate(
        direction = if_else(
          expression.trend > 0, "positive", "negative"
        )
      ) |>
      group_by(pert_name, direction) |>
      slice_head(n = 20),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = "red",
      "FALSE" = "black"
    ),
    guide = "none"
  ) +
  facet_wrap(
    ~pert_name,
    axes = "all", axis.labels = "margins",
    scales = "free"
  ) +
  labs(
    x = "Slope", y = "-log10(adj. p-value)"
  )
p

ggsave(
  file.path(plot_dir, "volcano_slopes_no_filtering.pdf"),
  p, width = 7, height = 4
)
```

## Compare with and without MDR filtering

```{r}
model_slope_comp <- power_inner_join(
  model_slopes |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(padj)
    ),
  model_mdr_included_slopes |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(padj)
    ),
  by = c("gene", "pert_name"),
  suffix = c("_filtered", "_unfiltered"),
  check = check_specs(
    unmatched_keys_left = "warn",
    unmatched_keys_right = "warn"
  )
)

p <- ggplot(
  model_slope_comp,
  aes(
    x = signed_p_unfiltered,
    y = signed_p_filtered
  )
) +
  geom_point(
    alpha = .6,
    shape = 16,
    color = "#aaaaaa"
  ) +
  geom_abline(
    linetype = "dashed",
    slope = 1, intercept = 0
  ) +
  geom_text_repel(
    data = \(x) bind_rows(
      x |> arrange(padj_filtered) |>
        filter(padj_filtered < 0.05) |>
        mutate(
          direction = if_else(
            direction_filtered == "negative", "negative", "positive"
          )
        ) |>
        group_by(pert_name, direction) |>
        slice_head(n = 5),
      x |> arrange(padj_unfiltered) |>
        filter(padj_unfiltered < 0.05) |>
        mutate(
          direction = if_else(
            direction_unfiltered == "negative", "negative", "positive"
          )
        ) |>
        group_by(pert_name, direction) |>
        slice_head(n = 5)
    ) |> distinct(),
    aes(label = gene),
    size = 1.5,
    max.overlaps = Inf,
    max.iter = 1e6,
    seed = 42,
    min.segment.length = 0
  ) +
  facet_wrap(
    ~pert_name,
    axes = "all", axis.labels = "margins",
    scales = "free"
  ) +
  labs(
    x = "Signed -log10(adj. p-value) without MDR filtering",
    y = "Signed -log10(adj. p-value) with MDR filtering"
  )

ggsave(
  file.path(plot_dir, "slope_p_value_comparison.pdf"),
  p, width = 10, height = 4
)
```


## Enrichment analysis


```{r}
database_abbreviations <- c(
  "REACTOME" = "R",
  "HALLMARK" = "H",
  "PID" = "P",
  "KEGG_MEDICUS" = "K",
  "GOBP" = "BP",
  "GOMF" = "MF",
  "GOCC" = "CC"
)

abbreviate_prefix <- function(strings, abbr_map = database_abbreviations) {
  # Create regex pattern from abbreviation names
  pattern <- paste0("^(", paste(names(abbr_map), collapse = "|"), ")_")

  # Replace matching prefixes with abbreviations
  str_replace(strings, pattern, function(x) {
    prefix <- str_remove(x, "_$")
    paste0(abbr_map[prefix], "_")
  })
}

all_gene_sets <- msigdbr(species = "Homo sapiens")
selected_gene_sets <- all_gene_sets %>%
  filter(
    gs_collection %in% c("H") |
      gs_subcollection %in% c(
        "CP:REACTOME",
        "CP:KEGG_MEDICUS"
      ),
    !str_detect(gs_name, coll("MEDICUS_PATHOGEN")),
    !str_detect(gs_name, coll("MEDICUS_VARIANT"))
  ) %>%
  mutate(
    database = str_extract(gs_name, paste0("^(", paste(names(database_abbreviations), collapse = "|"), ")")),
    pathway = abbreviate_prefix(gs_name)
  )

selected_gene_sets_list <- selected_gene_sets %>%
  group_nest(gs_collection, gs_subcollection, gs_name, database, pathway) %>%
  mutate(
    gs = map(
      data,
      \(x) unique(x$gene_symbol)
    ) %>%
      set_names(pathway)
  )

fora_input_raw <- tribble(
  ~model_type, ~de_type, ~data,
  "linear", "slopes", model_slopes |>
    filter(padj < 0.05) |>
    arrange(p.value) |>
    group_by(term = pert_name, direction) |>
    slice_head(n = 500),
  "linear", "pairwise", model_res_pairwise |>
    filter(padj < 0.05) |>
    arrange(p.value) |>
    group_by(term = contrast, direction) |>
    slice_head(n = 500),
  # "rank", "slopes", model_slopes_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = pert_name, direction) |>
  #   slice_head(n = 500),
  # "rank", "pairwise", model_res_pairwise_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = contrast, direction) |>
  #   slice_head(n = 500)
) |>
  mutate(
    across(data, \(x) map(x, ungroup))
  ) |>
  unnest(data) |>
  nest(.by = c(model_type, de_type, term, direction))

gene_universe <- names(eligible_genes)[eligible_genes]

fora_res_raw <- fora_input_raw %>%
  mutate(
    res = map(
      data,
      \(x) {
        withr::with_seed(42, fora(
          selected_gene_sets_list$gs,
          x$gene,
          universe = gene_universe
        ))
      }
    ),
    res_reduced = future_pmap(
      list(res, data),
      \(res, data) {
        collapsePathwaysORA(
          filter(res, padj < .05),
          selected_gene_sets_list$gs,
          data$gene,
          gene_universe
        )
      },
      .progress = TRUE
    )
  )

fora_res_long <- fora_res_raw %>%
  transmute(
    model_type, de_type, term, direction,
    res = map2(
      res, res_reduced,
      \(full, reduced) {
        mutate(
          full,
          is_main = pathway %in% reduced$mainPathways,
          parent_pathway = reduced$parentPathways[pathway]
        )
      }
    )
  ) %>%
  unnest(res) |>
  mutate(
    signed_padj = if_else(
      # multiplied by -1
      direction == "negative", 1, -1
    ) * log10(padj)
  )

```

```{r}
fora_res_top <- fora_res_long |>
  filter(padj < 0.05) |>
  arrange(padj, pval) |>
  group_by(model_type, de_type, term, direction) |>
  slice_head(n = 20) |>
  ungroup() |>
  arrange(padj) |>
  mutate(
    padj_cut = cut(
      padj,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("<0.001", "<0.01", "<0.05", "ns")
    ) |>
      fct_rev(),
    logEnrichment = log10(foldEnrichment) * if_else(direction == "negative", -1, 1),
    pathway_display = str_replace_all(pathway, coll("_"), " ")
  )


ps <- group_nest(
  fora_res_top, model_type, term
) |>
  mutate(
    p = map(
      data,
      \(df) {
        abs_max_est_perc <- max(abs(quantile(df$signed_padj, c(0.025, .975))), na.rm = TRUE)
        ggplot(
          df |>
            mutate(across(pathway_display, \(x) fct_reorder(x, logEnrichment))),
          aes(
            x = logEnrichment,
            y = pathway_display
          )
        ) +
          geom_vline(xintercept = 0, linetype = "dashed") +
          geom_point(
            aes(
              fill = signed_padj,
              size = padj_cut
            ),
            shape = 21,
            color = "black"
          ) +
          paletteer::scale_fill_paletteer_c(
            "ggthemes::Orange-Blue-White Diverging",
            limits = c(-abs_max_est_perc, abs_max_est_perc),
            oob = scales::squish
          ) +
          labs(
            x = "log10(Fold Enrichment)", y = NULL,
            fill = "log10(adj. p)",
            size = "Adjusted p"
          ) +
          envalysis::theme_publish()
          # scale_x_log10()
      }
    )
  )

dir.create(
  file.path(plot_dir, "enrichment"),
  showWarnings = FALSE, recursive = TRUE
)
pwalk(
  ps,
  \(term, p, model_type, ...) {
    ggsave(
      file.path(plot_dir, "enrichment", paste0(model_type, "_", term, "_enrichment.pdf")),
      p, width = 12, height = 7
    )
  }
)
```

### FGSEA

```{r}

fgsea_input_raw <- tribble(
  ~model_type, ~de_type, ~data,
  "linear", "slopes", model_slopes |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(padj)
    ) |>
    group_by(term = pert_name) |>
    summarize(
      ranks = list(set_names(signed_p, gene)),
      .groups = "drop"
    ),
  "linear", "pairwise", model_res_pairwise |>
    mutate(
      signed_p = if_else(
        direction == "negative", 1, -1
      ) * log10(padj)
    ) |>
    group_by(term = contrast) |>
    summarize(
      ranks = list(set_names(signed_p, gene)),
      .groups = "drop"
    )
  # "rank", "slopes", model_slopes_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = pert_name, direction) |>
  #   slice_head(n = 500),
  # "rank", "pairwise", model_res_pairwise_rank |>
  #   filter(padj < 0.05) |>
  #   arrange(p.value) |>
  #   group_by(term = contrast, direction) |>
  #   slice_head(n = 500)
) |>
  mutate(
    across(data, \(x) map(x, ungroup))
  ) |>
  unnest(data)

gene_universe <- names(eligible_genes)[eligible_genes]

fgsea_res_raw <- fgsea_input_raw %>%
  mutate(
    res = map(
      ranks,
      \(x) {
        withr::with_seed(42, fgseaMultilevel(
          selected_gene_sets_list$gs,
          x
        ))
      }
    ),
    res_reduced = future_pmap(
      list(res, ranks),
      \(res, x) {
        collapsePathways(
          filter(res, padj < .05),
          selected_gene_sets_list$gs,
          x
        )
      },
      .progress = TRUE
    )
  )

fgsea_res_long <- fgsea_res_raw %>%
  transmute(
    model_type, de_type, term,
    res = map2(
      res, res_reduced,
      \(full, reduced) {
        mutate(
          full,
          is_main = pathway %in% reduced$mainPathways,
          parent_pathway = reduced$parentPathways[pathway]
        )
      }
    )
  ) %>%
  unnest(res) |>
  mutate(
    direction = if_else(NES < 0, "negative", "positive"),
    signed_p = if_else(
      direction == "negative", 1, -1
    ) * log10(padj)
  )


```


```{r}
fgsea_res_top <- fgsea_res_long |>
  filter(padj < 0.05) |>
  arrange(padj, pval) |>
  group_by(model_type, de_type, term, direction) |>
  slice_head(n = 20) |>
  ungroup() |>
  arrange(padj) |>
  mutate(
    padj_cut = cut(
      padj,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("<0.001", "<0.01", "<0.05", "ns")
    ) |>
      fct_rev(),
    pathway_display = str_replace_all(pathway, coll("_"), " ")
  )


ps <- group_nest(
  fgsea_res_top, model_type, term
) |>
  mutate(
    p = map(
      data,
      \(df) {
        abs_max_est_perc <- max(abs(quantile(df$signed_p, c(0.025, .975))), na.rm = TRUE)
        ggplot(
          df |>
            mutate(across(pathway_display, \(x) fct_reorder(str_trunc(x, 50,), NES))),
          aes(
            x = NES,
            y = pathway_display
          )
        ) +
          geom_vline(xintercept = 0, linetype = "dashed") +
          geom_point(
            aes(
              fill = signed_p,
              size = padj_cut
            ),
            shape = 21,
            color = "black"
          ) +
          paletteer::scale_fill_paletteer_c(
            "ggthemes::Orange-Blue-White Diverging",
            limits = c(-abs_max_est_perc, abs_max_est_perc),
            oob = scales::squish
          ) +
          labs(
            x = "Normalized enrichment score", y = NULL,
            fill = "log10(adj. p)",
            size = "Adjusted p"
          )
          # scale_x_log10()
      }
    )
  )

dir.create(
  file.path(plot_dir, "enrichment"),
  showWarnings = FALSE, recursive = TRUE
)
pwalk(
  ps,
  \(term, p, model_type, ...) {
    ggsave(
      file.path(plot_dir, "enrichment", paste0("fgsea_", model_type, "_", term, "_enrichment.pdf")),
      p, width = 8, height = 8
    )
  }
)
```


## Make Excel sheet

```{r}
format_p <- function(wb, sheet_name, col_name, table) {
  data_rows <- nrow(table) + 1  # +1 for header
  col_idx <- which(names(table) == col_name)

  conditionalFormatting(
    wb = wb,
    sheet = sheet_name,
    cols = col_idx,
    rows = 2:data_rows,  # Skip header row
    type = "expression",
    rule = paste0(
      "AND(NOT(ISBLANK(", openxlsx::int2col(col_idx), "2)),",
      "ISNUMBER(", openxlsx::int2col(col_idx), "2),",
      openxlsx::int2col(col_idx), "2<0.05)"
    ),
    style = createStyle(bgFill = "#FF9999")  # Light red background
  )
}

format_phenotype <- function(wb, sheet_name, col_name, table, percentile = .001) {
  data_rows <- nrow(table) + 1  # +1 for header
  col_idx <- which(names(table) == col_name)

  min_max <- quantile(table[[col_name]], probs = c(percentile, 1 - percentile), na.rm = TRUE) |>
    abs() |>
    max()

  conditionalFormatting(
    wb = wb,
    sheet = sheet_name,
    cols = col_idx,
    rows = 2:data_rows,  # Skip header row
    type = "colourScale",
    style = c("blue", "white", "red"),
    rule = c(-1, 0, 1) * min_max
  )
}

add_sheet <- function(wb, sheet_name, table) {
  addWorksheet(wb, sheet_name)
  # table <- select(table, where(negate(is.list)))
  writeDataTable(wb, sheet_name, table, tableStyle = "TableStyleLight1")
  walk(
    str_subset(names(table), "^padj"),
    \(col_name) format_p(wb, sheet_name, col_name, table)
  )
  walk(
    str_subset(names(table), "^estimate|^expression.trend|^foldEnrichment|^signed_padj"),
    \(col_name) format_phenotype(wb, sheet_name, col_name, table)
  )
  setColWidths(wb, sheet_name, cols = seq_len(ncol(table)), widths = "auto")
}

wb <- createWorkbook()

list(
  "Slopes" = model_slopes |>
    arrange(padj, p.value),
  "Pairwise" = model_res_pairwise |>
    select(-term) |>
    arrange(padj, p.value)
) |>
  iwalk(\(t, n) add_sheet(wb, n, t))

saveWorkbook(
  wb,
  file.path(result_dir, "cdk7_prism_expression_model.xlsx"),
  overwrite = TRUE
)

wb <- createWorkbook()

fora_res_long |>
  arrange(padj, pval) |>
  relocate(signed_padj, .after = padj) |>
  relocate(overlapGenes, .after = last_col()) |>
  group_by(term) |>
  group_walk(\(df, key) {
    add_sheet(wb, key$term, df)
  })

saveWorkbook(
  wb,
  file.path(result_dir, "cdk7_prism_expression_enrichment.xlsx"),
  overwrite = TRUE
)
```
