---
title: "DGE"
author: "Clemens Hug"
date: "2023-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)
library(synExtra)
library(qs)

synapser::synLogin()
syn <- synDownloader("~/data", .cache = TRUE)
```

## Reading bcbio count matrices

```{r wrangling_funcs}
barcode_well_map <- syn("syn12979100") %>%
  readxl::read_xlsx(sheet = "barcodes_trugrade_384_set1")
  # rename(Barcode = barcode, Well = well, Plate_ID = plate_id)

ensembl_gtf_file <- "Homo_sapiens.GRCh38.109.gtf.gz"
if (!file.exists(ensembl_gtf_file)) {
  download.file(
    "ftp://ftp.ensembl.org/pub/release-109/gtf/homo_sapiens/Homo_sapiens.GRCh38.109.gtf.gz",
    ensembl_gtf_file, method = "curl"
  )
}

ensembl_gtf <- rtracklayer::readGFF(ensembl_gtf_file) %>%
  filter(gene_biotype == "protein_coding") %>%
  distinct(
    ensembl_gene_id = gene_id,
    hgnc_symbol = gene_name,
    gene_biotype
  ) %>%
  drop_na(ensembl_gene_id)

prepare_barcode_counts <- function(barcode_count_file, barcode_map) {
  read_tsv(barcode_count_file, col_names = c("barcode", "count")) %>%
    mutate(
      barcode = str_replace(barcode, "^.*([ATCG]{6})$", "\\1")
    ) %>%
    left_join(
      barcode_well_map, by = "barcode"
    )
}

prepare_col_meta <- function(colmeta_file, barcode_map) {
  read_csv(colmeta_file, col_names = "barcode") %>%
    mutate(
      barcode = if (all(str_detect(barcode, fixed(":")))) str_split(barcode, fixed(":"), simplify = TRUE)[, 2] else barcode,
      col_idx = 1:n()
    ) %>%
    left_join(
      barcode_map %>%
        select(barcode, well),
      by = "barcode"
    )
}

prepare_row_meta <- function(rowmeta_file) {
  read_csv(rowmeta_file, col_names = "ensembl_gene_id") %>%
    mutate(row_idx = 1:n())
}

prepare_mtx_sparse <- function(mtx_file) {
  read_delim(
    mtx_file,
    delim = " ",
    skip = 3,
    col_names = c("row_idx", "col_idx", "count"),
    col_types = "iii"
  )
}

prepare_counts <- function(mtx, col_meta, row_meta, gene_mapping) {
  # browser()
  mtx %>%
    left_join(col_meta, by = "col_idx") %>%
    left_join(row_meta, by = "row_idx") %>%
    select(well, ensembl_gene_id, count) %>%
    left_join(
      gene_mapping, by = "ensembl_gene_id"
    )
}

```
Reading DGE metadata and counts from synapse

```{r load}
meta <- syn("syn51181405") %>%
  read_csv() %>%
  mutate(
    well_384 = str_replace(well_384, "^([A-Z])([0-9])$", "\\10\\2")
  ) %>%
  mutate(
    experiment_type = case_when(
      is.na(agent) ~ "empty",
      agent == "density" ~ "density",
      TRUE ~ "compound"
    )
  )

raw_counts <- prepare_counts(
  prepare_mtx_sparse(syn("syn51181399")),
  col_meta = prepare_col_meta(syn("syn51181400"), barcode_well_map),
  row_meta = prepare_row_meta(syn("syn51181404")),
  ensembl_gtf
) %>%
  select(-gene_biotype)

```
## Count plots

```{r}
library(ggbeeswarm)

well_counts <- raw_counts %>%
  group_by(well) %>%
  summarize(count = sum(count), .groups = "drop") %>%
    left_join(
    meta,
    by = c("well" = "well_384")
  )

p <- well_counts %>%
  drop_na(source_96) %>%
  ggplot(aes(source_96, count, color = cell_line, shape = experiment_type)) +
    geom_quasirandom() +
    scale_y_log10() +
    geom_hline(yintercept = 5e4) +
    scale_shape_manual(values = c(density = 3, compound = 16, empty = 4)) +
    labs(x = "Plate", y = "Total count per well", color = "Cell line", shape = "Experiment type")

dir.create("plots/qc", recursive = TRUE)
ggsave(
  "plots/qc/well_counts_beeswarm.pdf", p, width = 10, height = 7
)
```

Plotting total gene counts per well against the number of cells seeded into
each well. Using empty wells as 0 density.

```{r}
p <- well_counts %>%
  filter(experiment_type %in% c("density", "empty")) %>%
  mutate(
    density = if_else(
      experiment_type == "density", concentration, 0
    )
  ) %>%
  ggplot(aes(density, count)) +
    geom_jitter(width = 1000) +
    labs(x = "Seeding density", y = "Total count per well")

ggsave(
  "plots/qc/well_counts_vs_density.pdf", p, width = 6, height = 5
)
```

## DESeq2

Preparing counts and metadata for DESeq2 analysis.

Excluding any wells with fewer than 5e4 counts.

```{r}
library(DESeq2)

de_meta <- well_counts %>%
  filter(
    experiment_type != "empty",
    count > 5e4
  ) %>%
  drop_na(source_well_96) %>%
  mutate(
    condition = paste(cell_line, agent, concentration, timepoint, sep = "_")
  )

de_counts <- raw_counts %>%
  filter(well %in% de_meta$well) %>%
  select(
    well, ensembl_gene_id, count
  ) %>%
  pivot_wider(
    names_from = well, values_from = count, values_fill = 0
  )

dir.create("deseq")
write_csv(
  de_counts, "deseq/de_counts.csv.gz"
)
write_csv(
  de_meta, "deseq/de_meta.csv.gz"
)
```

Only consider genes that have at least 5 counts in more than 10 samples to
speed up analysis. Doesn't have impact on outcome because DESeq2 would exclude
these anyways, but take longer.

```{r}
genes_to_keep <- raw_counts %>%
  group_by(ensembl_gene_id) %>%
  summarize(n_over_five = sum(count > 5), .groups = "drop") %>%
  filter(n_over_five >= 10)

de <- DESeqDataSetFromMatrix(
  de_counts %>%
    column_to_rownames("ensembl_gene_id") %>%
    as.matrix() %>% {
      .[genes_to_keep$ensembl_gene_id, de_meta$well]
    },
  de_meta,
  design = ~condition
)
qsave(
  de, "deseq/de_input.qs"
)
```

For exploratory analysis (PCA etc), use only genes that are differentially
expressed in *any* condition. To find these genes, use Likelihood-ratio test (LRT)
from DESeq2, comparing a model with a single term for every condition (~condition) 
against an intercept-only null model (~1).

```{r}
des <- DESeq(
  de, test = "LRT", reduced = ~1,
  parallel = TRUE, BPPARAM = BiocParallel::MulticoreParam(workers = 4)
)
dir.create("qc")
qsave(
  des, "qc/deseq_lrt_res.qs"
)

res_lrt <- results(des, tidy = TRUE)


```

## Exploratory analysis

Before PCA, do variance stabilizing transformation improving the distribution
properties of the counts, similar to, but superior to, to doing log10(x + 1).

Only use genes that have adjusted p-value of <0.01 from LRT before.

```{r}
vsd <- varianceStabilizingTransformation(de)
counts_vst <- assay(vsd) %>%
  as_tibble(rownames = "ensembl_gene_id")

genes_for_pca <- res_lrt %>%
  filter(padj < 0.01) %>%
  pull("row")

pca_raw <- prcomp(assay(vsd)[genes_for_pca, ])

pca_tidy <- broom::tidy(pca_raw, matrix = "rotation")

pca_var_tidy <- broom::tidy(pca_raw, matrix = "eigenvalues")
```

Plot all combinations of components PC1 to PC4 against each other.

```{r}
ps <- combn(1:4, 2, simplify = FALSE) %>% {
  tibble(component_1 = map_int(., 1), component_2 = map_int(., 2))
} %>%
  rowwise() %>%
  mutate(
    p = list(
      ggplot(
          pca_tidy %>%
            filter(PC %in% c(component_1, component_2)) %>%
            pivot_wider(names_from = PC, values_from = value, names_prefix = "PC_") %>%
            left_join(meta, by = c("column" = "well_384")),
          aes_string(paste0("PC_", component_1), paste0("PC_", component_2), color = "cell_line", shape = "source_96")
        ) +
          geom_point() +
          labs(
            x = paste0("PC ", component_1, " (", signif(filter(pca_var_tidy, PC == component_1)$percent * 100, digits = 2), "%)"),
            y = paste0("PC ", component_2, " (", signif(filter(pca_var_tidy, PC == component_2)$percent * 100, digits = 2), "%)"),
            shape = "Plate",
            color = "Cell line"
          )
    )
  )

library(patchwork)
ps_combined <- wrap_plots(
  ps$p, guides = "collect"
)

ggsave(
  "plots/qc/vsn_counts_pca.pdf", ps_combined, width = 14, height = 10
)
```

For good measure, also attempting UMAP.

```{r}
set.seed(42)
umap_raw <- uwot::umap(
  assay(vsd)[genes_for_pca, ] %>%
    t(),
  n_neighbors = 5,
  init = "spectral",
  scale = "none"
)

umap_tidy <- umap_raw %>%
  as_tibble(rownames = "well") %>%
  left_join(meta, by = c("well" = "well_384"))

p <- umap_tidy %>%
  ggplot(
    aes(V1, V2, color = cell_line, shape = source_96)
  ) +
    geom_point()
ggsave(
  "plots/qc/vsn_counts_umap.pdf", p, width = 8, height = 6
)
```

## Old DGE experiments

Wrangle old DGE experiments, finding conditions that overlap that we can
compare to the new results.

```{r old_samples}
library(fst)
lsp_compound_mapping <- syn("syn26260389") %>%
  read_fst()
lsp_compound_names <- syn("syn26260347") %>%
  read_fst(as.data.table = TRUE) %>% {
    .[
      ,
      head(.SD, n = 1),
      keyby = .(lspci_id)
    ]
  }

raw_counts <- prepare_counts(
  prepare_mtx_sparse(syn("syn51181399")),
  col_meta = prepare_col_meta(syn("syn51181400"), barcode_well_map),
  row_meta = prepare_row_meta(syn("syn51181404")),
  ensembl_gtf
) %>%
  select(-gene_biotype)


old_samples_syn <- tribble(
  ~experiment, ~file_type, ~synapse_id,
  "dge1", "mtx", "syn18143723",
  "dge1", "colnames", "syn18143725",
  "dge1", "rownames", "syn18143724",
  "dge2", "mtx", "syn17103650",
  "dge2", "colnames", "syn17103652",
  "dge2", "rownames", "syn17103651",
  "dge3", "mtx", "syn25764600",
  "dge3", "colnames", "syn25764602",
  "dge3", "rownames", "syn25764604"
) %>%
  mutate(
    file = map_chr(synapse_id, syn)
  )

old_samples_raw_counts <- old_samples_syn %>%
  pivot_wider(
    names_from = file_type, values_from = c(synapse_id, file)
  ) %>%
  rowwise() %>%
  mutate(
    counts = prepare_counts(
      prepare_mtx_sparse(file_mtx),
      col_meta = prepare_col_meta(file_colnames, barcode_well_map),
      row_meta = prepare_row_meta(file_rownames),
      ensembl_gtf
    ) %>%
      select(-gene_biotype) %>%
      list()
  ) %>%
  ungroup()

old_samples_raw_counts_long <- old_samples_raw_counts %>%
  select(experiment, counts) %>%
  unnest(counts)

fwrite(old_samples_raw_counts_long, "deseq/old_samples_raw_counts_long.csv.gz")

old_sample_metas <- tribble(
  ~experiment, ~meta,
  "dge1", syn("syn11947020") %>%
    fread() %>%
    mutate(
      `Fluid name` = recode(
        `Fluid name`,
        `Drug control` = "DMSO"
      )
    ) %>%
    select(well = `Dispensed well`, agent = `Fluid name`, concentration = `Concentration`),
  "dge2", syn("syn17114439") %>%
    fread() %>%
    mutate(
      `Fluid name` = if_else(`Fluid name` == "", "DMSO", `Fluid name`)
    ) %>%
    transmute(well = `Dispensed well`, agent = `Fluid name`, concentration = as.character(`Concentration`)),
  "dge3", syn("syn25925490") %>%
    fread() %>%
    left_join(
      lsp_compound_mapping %>%
        select(lspci_id, external_id),
      by = c("agent" = "external_id")
    ) %>%
    left_join(
      lsp_compound_names,
      by = c("lspci_id")
    ) %>%
    transmute(well, agent = if_else(agent == "DMSO", "DMSO", name), cell_line, concentration = as.character(concentration), role, timepoint)
) %>%
  unnest(meta)


fwrite(old_sample_metas, "deseq/old_samples_metas.csv.gz")
```

Manual inspection reveals that three drug treatments overlap.
Only consider those.

```{r}
old_samples_de_meta <- old_sample_metas %>%
  mutate(
    agent = recode(
      agent,
      `KW-2449` = "KW2449",
      `kw2449` = "KW2449"
    )
  ) %>%
  filter(
    agent %in% c(
      "Baricitinib",
      "DMSO",
      "KW2449",
      "Ruxolitinib"
    )
  )

fwrite(old_samples_de_meta, "deseq/old_samples_de_meta.csv.gz")

```

